---
title: "Stats 306: Lab 4"
subtitle: "Data Manipulation"
author: "Yilei Zhang"
output: 
  slidy_presentation:
    incremental: true
runtime: shiny_prerendered
---

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# This just checks if students need to install some packages that they might 
# not have.
if (!require(gapminder)) install.packages("gapminder", repos = "http://cran.us.r-project.org")

```

```{r setup, echo=FALSE, include=FALSE}
library(learnr)     # For interactive exercises
library(tidyverse)  # For ggplot2, dplyr
library(gapminder)
```

## Logistics

> - HW2 is due on Oct 2.



## 

> - Review of Quiz 2
> - Directory Commands


## Today's Content

> - mutate, transmute, rename
> - group_by, summarise
> - pipe operator


## Mutate

We can use `mutate` to add new variables and preserve existing ones. We will look at a few examples. Here we use the dataset iris.

```{r iris}
head(iris)
```


## Basics of Mutate

> - Create a Sepal Area column that equals to `Sepal Width * Sepal Length`

```{r}
iris1 <- mutate(iris, Sepal.Area = Sepal.Width * Sepal.Length)
head(iris1)
```

> - Make this new column before Sepal Length

```{r}
iris1 <- mutate(iris, Sepal.Area = Sepal.Width * Sepal.Length,
                      .before = Sepal.Length)
head(iris1)
```

There are more useful arguments:

> - `.keep = c("all", "used", "unused", "none")`
> - `.before` and `.after` - optionally specifies the position of the new column

```{r mutate, exercise=T}
?mutate
```


## Creating Multiple Columns

You can create multiple columns, some of which may depend on those being created at the same time.

```{r}
## Create Sepal Area column as Sepal Width * Sepal Legth
## Update this area by scaling it by a factor of 10
iris1 <- mutate(iris, Sepal.Area = Sepal.Width * Sepal.Length,
               Scaled.Sepal.Area = Sepal.Area*10)
head(iris1)
```

If we only want to keep the new columns and those columns used to create the new columns in the resulting dataset:

```{r}
iris1 <- mutate(iris, Sepal.Area = Sepal.Width * Sepal.Length,
               Scaled.Sepal.Area = Sepal.Area*10,
               .keep = "used")
head(iris1)
```

*Note*: We can use mutate to delete a column by setting the column name to NULL

```{r}
## Remove Species column
iris1 <- mutate(iris, Species = NULL)
head(iris1)
```

## Using Functions

> - We can use function `row_number()` to get row id.
> - Use function `toupper()` to convert characters to upper cases.

```{r}
iris1 <- mutate(iris, ID = row_number(),
                Species.Upper = toupper(Species))
head(iris1)
```

> - Use `lag()` and `lead()` to get previous and later rows respectively.

```{r}
iris1 <- mutate(iris, previous = lag(Sepal.Length),
                      future = lead(Sepal.Length))
head(iris1)
```

> - Use `case_when()` to create a conditional variable, `Size`: if `Sepal Length` is less than the average, assign the value "small", if it is greater than the average, assign "large", else assign "medium".

```{r}
iris1 <- mutate(iris, Size = case_when(
                 Sepal.Length < mean(Sepal.Length) ~ "small",
                 Sepal.Length > mean(Sepal.Length) ~ "large",
                   TRUE ~ "medium")
                   )
head(iris1)
```

Like an if statement, the arguments are evaluated in order, so you must
proceed from the most specific to the most general. For example, this won't work:

```{r}
x <- 1:50
case_when(
  TRUE ~ as.character(x),
  x %%  5 == 0 ~ "fizz",
  x %%  7 == 0 ~ "buzz",
  x %% 35 == 0 ~ "fizz buzz"
)
```

## Mutate Multiple Columns

```{r}
iris1 <- mutate(iris, across(starts_with("Sepal"), ~ . / 10)) 
head(iris1)
```

Breaking down the syntax above:

- across: specifies a set of columns to mutate
- ~ indicates the start of a formula to apply across several columns
- . dummy variable standing in for each column


## transmute

`transmute` works exactly like `mutate`, the only difference being that only newly created columns will be retained in our new dataset (tibble). It is exactly the same as setting .keep = "none".

We can create Sepal Area and Scaled Sepal Area columns as a separate dataset

```{r}
iris2 <- transmute(iris, Sepal.Area = Sepal.Width * Sepal.Length,
               Scaled.Sepal.Area = Sepal.Area*10)
head(iris2)
```

Alternatively, we can create Sepal Area and Scaled Sepal Area columns as a separate dataset using mutate

```{r}
iris2 <- mutate(iris, Sepal.Area = Sepal.Width * Sepal.Length,
               Scaled.Sepal.Area = Sepal.Area*10,
               .keep = "none")
head(iris2)
```


## Rename

We use `rename` to rename existing column names.

dots(.) are not recommended for variable names, we want to rename all columns names in the iris dataset by replacing . with _

```{r}
iris2 <- rename(iris, sepal_length = Sepal.Length,
                      sepal_width = Sepal.Width,
                      petal_length = Petal.Length,
                      petal_width = Petal.Width)
head(iris2)
```

We have many repetitive codes above. We can also use `rename_with()` to apply a (string) function to all columns by default or a specific subset.

```{r}
iris2 <- rename_with(iris, ~ str_replace(., pattern="\\.", replacement = "_"))
head(iris2)
```

> - For all the columns in iris, we apply the function `str_replace()`.
> - We use "\\\\" to escape.

## summarize and group_by

`summarise` is used to compute a new dataset (tibble) by using aggregation functions. It is usually useful when used with `group_by`, as this helps to compute summaries for each category of a column of a dataset.

Some useful aggregation functions include:

> - Center: mean(), median()
> - Spread: sd(), IQR(), mad()
> - Range: `min(), max(), quantile()
> - Position: first(), last(), nth()
> - Count: n(), n_distinct()
> - Logical: any(), all()


## Summarise for entire dataset

To find the mean of Sepal Width and Sepla Length for the iris dataset:

```{r}
summarise(iris, mean_sep_width = mean(Sepal.Width), 
          mean_sep_length = mean(Sepal.Length)
          )

```

Question: What does the code below do?

```{r}
summarise(iris, n = n(), 
             unique= n_distinct(Species),
             first = first(Species), 
             last = last(Species),
             `30st` = nth(Species, 30))
```


## Summarise by groups

We first group the iris dataset by Species.

```{r}
by_group <- group_by(iris, Species)

```

Then take the mean for Sepal Length and Width within each group:

```{r}
summarise(by_group, mean_sep_length = mean(Sepal.Length),
          mean_sep_width = mean(Sepal.Width))
```


## The Pipe Operator

As you probably saw from the previous example, it is extraneous to be storing intermediate steps while completing a chain of tasks. The pipe operator (%>%) helps up eliminate this hassle.

Consider how the pipe (%>%) can help eliminate such hassle in the following sequence (chain) of tasks:

> - Filter the iris dataset for Petal Length greater than 1 and Sepal Width less than 3.3
> - Select Sepal Length, Sepal Width and Species
> - Rename the columns of Sepal Width and Sepal Length by replacing dot(.) with underscore(_)
> - Convert the Species column to uppercase
> - Group the resulting dataset by species and compute the 50th percentile and the mean for each group

Note: An easier way to understand the pipe (%>%) operator is reading it as "and then"

```{r}
## Using piping to accomplish the above task:
iris %>%                                                ## then
    filter(Petal.Length > 1 & Sepal.Width < 3.3) %>%    ## and then
    select(Sepal.Length, Sepal.Width, Species) %>%      ## and then
    rename(sepal_length = Sepal.Length,
           sepal_width = Sepal.Width,) %>%              ## and then
    mutate(Species = toupper(Species)) %>%              ## and then
    group_by(Species) %>%                               ## and then
    summarise(mean_sep_length = mean(sepal_length),
              mean_sep_width = mean(sepal_width),
              `50th_tile_sep_len` = quantile(sepal_length, 0.5),
              `50th_tile_sep_width` = quantile(sepal_width, 0.5)
              )
```


## Exercises

We use the dataset `gapminder` for the exercises:

```{r}
head(gapminder)
```

## Question 1

Complete the following tasks:

> - subset for European and Asian countries excluding Japan, Norway and Israel
> - scale the population by dividing by 1000
> - select all columns aside from those ending with "y"
> - sort the resulting tibble by continent ascending and gdp per capita in descending order
> - Retain only rows 5 to 20

What is the frist row in the resulting dataset?

```{r question1, exercise = T}

```


## Question 2

Complete the following tasks:

> - group the dataset by year and continent
> - compute the total population (hint: use sum()), mean life expectancy and count the number of cases for each group
> - make a scatterplot of year by total population continent. The plots should be separate for each continent.

```{r question2, exercise = T}

```

## Question 3

- subset for only the year 2007
- group by continent
- compute the mean life expectancy and count the number of cases for each group
- create:
  + a bar chart for continent, where the bars are horizontal than vertical
  + create a pie chart for the counting variable
    - fill by continent
    - add texts superimposing the counts of each continent on the pie chart [hint]:   geom_text or geom_label
    - move the legend to the bottom position than the right position

```{r question3, exercise = T}

```

## Solutions

Question 1:

> - subset for European and Asian countries excluding Japan, Norway and Israel
> - scale the population by dividing by 1000
> - select all columns aside from those ending with "y"
> - sort the resulting tibble by continent ascending and gdp per capita in descending order
> - Retain only rows 5 to 20

What is the frist row in the resulting dataset?

```{r solution1, exercise = T}
gapminder %>% ## then
    filter(continent %in% c("Europe", "Asia"), !country %in% c("Japan", "Norway", "Israel")) %>%
    mutate(`scaled pop` = pop/1e3) %>%
    select(!ends_with("y")) %>%
    arrange(continent, desc(gdpPercap)) %>%
    slice(5:20)
```


Question 2:

Complete the following tasks:

> - group the dataset by year and continent
> - compute the total population (hint: use sum()), mean life expectancy and count the number of cases for each group
> - make a scatterplot of year by total population continent. The plots should be separate for each continent.

```{r solution2, exercise = T}
gapminder %>%
   group_by(year, continent) %>%
   summarise(total_pop = sum(pop),
             mean_life_expt = mean(lifeExp),
             counts = n()) %>%
   ggplot(aes(x = year, y = total_pop, color = continent)) +
   geom_point() +
   facet_wrap(~ continent) +
   theme_bw() ## add black and white background
```


Question 3:

- subset for only the year 2007
- group by continent
- compute the mean life expectancy and count the number of cases for each group
- create:
  + a bar chart for continent, where the bars are horizontal than vertical
  + create a pie chart for the counting variable
    - fill by continent
    - add texts superimposing the counts of each continent on the pie chart [hint]:   geom_text or geom_label
    - move the legend to the bottom position than the right position

```{r solution3, exercise = T}
df <- gapminder %>%
        filter(year == 2007) %>%
        group_by(continent) %>%
        summarise(mean_life_expt = mean(lifeExp),
                  counts = n())

## (1) making the bar chart
df %>% ggplot() +
         geom_col(mapping = aes(x = continent, y = mean_life_expt, fill = continent)) +
         coord_flip() + ## making bars horizontal 
         theme_bw()  ## black and white theme

## (2) making pie chart
df %>% ggplot(mapping = aes(x = 1, y = counts, fill = continent)) +
         geom_col(width = 1, color = "white") +
         coord_polar(theta = "y") +
         ## adding labels
         geom_label(mapping = aes(label = counts), ## alternatively use geom_text which has no border around the labels
                    color = "black",               ## border colr
                    position = position_stack(vjust = 0.5), 
                    show.legend = FALSE) +
         theme_void() +                            ## clears backgrond
         theme(legend.position = "bottom")         ## move legend to bottom
```

