---
title: "Stats 306: Lab 12"
author: "Your Name"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

## Set up

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# This just checks if students need to install some packages that they might 
# not have.

if (!require(bench))
  install.packages("bench", repos = "http://cran.us.r-project.org")

if (!require(parallel))
  install.packages("parallel", repos = "http://cran.us.r-project.org")

if (!require(doParallel))
  install.packages("doParallel", repos = "http://cran.us.r-project.org")

if (!require(foreach))
  install.packages("foreach", repos = "http://cran.us.r-project.org")
```

```{r setup, eval=TRUE, include=TRUE, message=FALSE, echo=TRUE}
library(learnr)     # For interactive exercises
library(purrr)
library(dplyr)
library(ggplot2)
library(bench)
library(parallel)
library(foreach)
library(doParallel)
```

```{r, include=FALSE, message=FALSE, echo=FALSE}
tutorial_options(exercise.reveal_solution=FALSE)
```

## Logistics

- HW 5
- End of the semester logistics:
    - After the lab today, one more lab next week. 
    - There will be a review session, probably Friday Dec 9, around 5:30 pm. To be scheduled. 
- Last quiz this weekend (debugging & performance)
- Final Exam

## Today's Content

- Debugging Review
- Performance Evaluation in `R` 
- Brief intro to parallelization in `R` 

## Debugging Review 

As you've likely seen many times this semester, incorrect `R` code can sometimes result in an *Error* message. 
When this happens, we say that `R` *throws* an error. 

**Example:**

```{r varianceFun, error=TRUE}
variance(rnorm(100))
```

Often the error message will help us find what is wrong with our code. This is the case above, where the error message tells us that the function `variance` could not be found. 
Sometimes, however, the error message is not as informative. 

**Example:**

```{r lmMean, error=TRUE}
lm(mean)
```

As outlined in the [Advanced R](https://adv-r.hadley.nz/debugging.html) textbook, a good strategy to help identifying and fixing errors is the following: 

> (1) **Google!**: 
> (2) **Make it Repeatable**
> (3) **Figure out where it is**
> (4) **Fix it and test it** 

**Example: Make it Repeatable.** Let's say that we want to create a random variable in the following 
way: 

- Sample $X_i \sim N(3, 1)$ distribution, for $i \in 1:100$
- Set $Y_i = log(X_i)$ for each $i$, 
- If $Y_i < 0$, set $Y_i = 0$.

We implement this strategy for sampling $Y$ below. 

```{r}
x <- rnorm(100, mean = 3, sd = 1)
log_x <- log(x)
for (i in 1:length(x)) {
  if (log_x[i] < 0) {
    log_x[i] <- 0
  }
}

hist(log_x)
```

Note that sometimes this results in an error, but usually it doesn't! This makes 
it really difficult to debug! 
For this particular problem, it might be easy to find the bug, but this example illustrates how getting a reproducible error (step 2) can help us find the bug. 

### Exercise:

Set the reproducible seed number so that the code below results in an error. 

```{r setSeed, exercise=TRUE}
set.seed(1)  # Change 1 to any other number until the code has an error. 
x <- rnorm(100, mean = 3, sd = 1)
log_x <- log(x)
for (i in 1:length(x)) {
  if (log_x[i] < 0) {
    log_x[i] <- 0
  }
}

hist(log_x)
```

```{r setSeed-solution, error=TRUE}
# There are many possible solutions, but here is one of them: 
set.seed(8)
x <- rnorm(100, mean = 3, sd = 1)
log_x <- log(x)
for (i in 1:length(x)) {
  if (log_x[i] < 0) {
    log_x[i] <- 0
  }
}

# One clever way to do this is by setting the seed in a for loop, and when it 
# results in an error, you'll have the seed number saved that broke the code.

hist(log_x)
```

Now that the error is reproducible, it should be easier to debug. 

### Exercise: 

Using `set.seed(8)`, find the first value of $i$ in the for-loop that causes an error. What's special about this value of $i$? 

```{r whichI, exercise=TRUE}
set.seed(8)  
x <- rnorm(100, mean = 3, sd = 1)
log_x <- log(x)
for (i in 1:length(x)) {
  if (log_x[i] < 0) {
    log_x[i] <- 0
  }
}
```

```{r whichI-solution}
# Because the code throws an error, the for loop stops. Thus, the value of i 
# that causes the error is available after the for-loop finishes. 
# 
# Another solution could be to print each value of i in the for-loop. 
set.seed(8)  
x <- rnorm(100, mean = 3, sd = 1)
log_x <- log(x)
for (i in 1:length(x)) {
  if (log_x[i] < 0) {
    log_x[i] <- 0
  }
}

cat("x[i] = ", x[i])  
# x[i] is negative, so taking the log results in NaN, which causes an error in 
# the if statement, because we can't compare missing values to a number. 
```

### Optional Exercise

Save this exercise for the end, if there is time. 

Open the file `lab12/plotMin_bug.R`, and use any debugging tool to identify potetential bug(s) in the file and fix them. 
This script will plot a function and it's minimum value.

## Performance Evaluation in `R`

`R` is a really good programming language for Statistics and Data Science, but it does have some flaws. In the introduction to [Advanced R](https://adv-r.hadley.nz/introduction.html), Hadley Wickham (a big advocate of the `R` programming language) notes some potential issues with `R`, including the following: 

> Much of the `R` code you’ll see in the wild is written in haste to solve a pressing problem. As a result, code is not very elegant, fast, or easy to understand.
> `R` is not a particularly fast programming language, and poorly written `R` code can be terribly slow. `R` is also a profligate user of memory.

Performance evaluation tools can help us improve our `R` code and avoid these issues. 
Specifically, in lecture we discussed *profiling* and *(micro)benchmarking* as useful tools to help evaluate the performance of our code. 

**Exercise:** 

```{r profileVSbench}
question(
  "What is the difference between profiling and benchmarking?",
  answer("There is no significant difference, both functions do the same thing."),
  answer("Profiling is used to get an accurate estimate of how long a function will take, whereas benchmarking is used to determine how long each step of a function (or program) takes."),
  answer("Profiling is used to determine how long each step of a function (or program) takes, whereas benchmarking is used to get an accurate estimate of how long a function will take.", correct = TRUE),
  answer("Profiling compares the speed of a function to other functions, and benchmarking does not.")
)
```


### **Exercise: calculating the mean of a vector** 

Use the function `bench::mark` to determine which is the fastest way to compute the mean of a vector. 

```{r timeMean, exercise = TRUE}
set.seed(234532)
x <- rnorm(n = 100000, mean = 1.5, sd = 2)

mean1 <- function(x) {
  mean(x)
}

mean2 <- function(x) {
  sum(x) / length(x)
}

mean3 <- function(x) {
  tmp <- 0
  for (i in 1:length(x)) {
    tmp <- tmp + x[i]
  }
  tmp / length(x)
}

```


```{r timeMean-solution}
set.seed(234532)
x <- rnorm(n = 100000, mean = 1.5, sd = 2)

mean1 <- function(x) {
  mean(x)
}

mean2 <- function(x) {
  sum(x) / length(x)
}

mean3 <- function(x) {
  tmp <- 0
  for (i in 1:length(x)) {
    tmp <- tmp + x[i]
  }
  tmp / length(x)
}

mark(mean1(x), mean2(x), mean3(x))
```

### Exercise: speeding up a for loop. 

One common approach to speeding up `R` code is by replacing for-loops with vectorized operations. See if you can write a vectorized version of the function below that is faster than the for loop, and demonstrate that you get the same result but faster using `bench::mark`. 

```{r slowFor, exercise = TRUE}
set.seed(623)
x <- rbeta(5000, 0.5, 1.4)

f1 <- function(x) {
  n <- length(x)
  num_gt_05 <- 0
  
  for (i in 1:n) {
    if (x[i] > 0.5) {
      num_gt_05 <- num_gt_05 + 1
    }
  }
  num_gt_05 / n
}

f2 <- function(x) {
  # TODO: Edit this function
}
```

```{r slowFor-hint-1}
# By default, the function bench::mark checks if the two functions give the same results. 
```

```{r slowFor-hint-2}
# What does sum(c(TRUE, TRUE, FALSE, FALSE, TRUE)) return? How might we take 
# advantage of this? 
```

```{r slowFor-solution}
set.seed(623)
x <- rbeta(5000, 0.5, 1.4)

f1 <- function(x) {
  n <- length(x)
  num_gt_05 <- 0
  
  for (i in 1:n) {
    if (x[i] > 0.5) {
      num_gt_05 <- num_gt_05 + 1
    }
  }
  num_gt_05 / n
}

f2 <- function(x) {
  # One possible solution: 
  sum(x > 0.5) / length(x)
}

bench::mark(f1(x), f2(x))
```


## Parallelization 

One way to speed up your code is by parallization. 
Most modern computers have processors with multiple *cores*. Each *core* is basically a "brain" of the computer, as it receives instructions and then performs the desired calculations that satisfy those instructions. 
When we normally run `R`, we are only running our commands on a single core. 
Some operations, however, can be computed in parallel, meaning that one operation can be performed independently of another operation. 
When this is the case, we can *parallelize* our code so that it runs on more cores than one; if done correctly, this speeds up the computation time of our code. 
For example, if we run a porperly parallized command on 4 cores instead of 1, this should run approximately 4 times faster! 

There are many tools to do parallization in `R`, but here we focus on only a few: 

- `parallel` package: `parLapply`
- `foreach` and `doParallel`

### `parallel` package

For more resources, see [this Umich Website](https://dept.stat.lsa.umich.edu/~jerrick/courses/stat701/notes/parallel.html).
As mentioned in the linked website, parallelization in `R` is easiest if you're using a Macbook or Linux computer. In section, we'll just demonstrate the parallelization techniques that work for all systems (including Windows). 

First, we can check how many cores our computer has with the command: 

```{r}
numCores <- detectCores()
numCores
```

Now, we create a *Socket Cluster*. When we make a Socket Cluster, we are starting new `R` processes on our computer, so this means that packages and variables that were created outside of the cluster will not be available to us. 

```{r}
cl <- makeCluster(numCores)
```

Now let's take the mean of each column of the `iris` dataset using all of our computer cores: 

```{r}
parSapply(cl, iris, mean, na.rm = TRUE)
```

Note that the code above is simply running an `sapply` in a parallelized fashion: each column of the dataset is an input for the function `mean`, and each calculation is being calculated on a different computer core at the same time! 

When we are done, we need to close the cluster so that we aren't wasting our computer's resources: 

```{r}
stopCluster(cl)
```

## `foreach` and `doParallel` package

Another great tool is the `foreach` and `doParallel` packages. These allow us to write for-loops that are computed in parallel: 

```{r}
registerDoParallel(numCores)  # setting up system for parallelization

normal_sqrt <- function(n) {
  res <- c() 
  for (i in 1:n) {
    res <- c(res, sqrt(i))
  }
  res
}

par_sqrt <- function(n) {
  foreach(i=1:n, .combine = c) %dopar% {
    sqrt(i)
  }
}

normal_sqrt(10)
par_sqrt(10)
```

If you were to time the two functions above, you'd likely see that the parallel version is actually slower. This is because there is some overhead when using `%dopar%` with `foreach`, and each operation is relatively quick. 
This fact highlights when parallelization is most effective: each iteration in a calculation is computationally intensive when compared to the parallelization overhead. 
