---
title: "Stats 306, Fall 2022, Homework 1"
author: "Your Name, Your Uniqname"
date: "Due Sept 18, 11:59PM EST"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Question 1 (5 points, 1 each per subproblem)

In this section, we will use git from the command line to learn more about
version control. If you have this document, you have probably already used git
to retrieve our class repository. If not, follow the directions on our Canvas
home page to get your own copy of the repository.

When you start RStudio, typically the *Terminal* pane will default to your home
directory. If you open the `stats306_fall_2022.rpoj` file, then the terminal
will open in the repository's directory. You verify this using the print working
directory command,  `pwd`. You may need to use the change directory command
(`cd`) to move into your repository. If you forget the name of your repository,
use `ls` to list files.

### Setting up git

In the first lecture (`lectures/lecture00.Rmd` in this repository), there are
directions to set up your git configuration to include your name and email
address. Using the *Terminal* pane, use these commands to set up your name
and email address.

The `config` subcommand also lets you retrieve settings. Use `git config` with
the `get` option (recall: how are *optional* arguments passed to the command
line arguments) to retrieve your `user.name` and `user.email`.

Put a copy of your session here:

```

```
### Using the log

Every commit in git gets its own identifier and log message. Using the `git log`
command you can see the entire log. Since this is quite long, it will be
"paginated". You can scroll up and down using arrow keys or the space bar to
move pages. Hit 'q' to quit.

For the most recent commit, fill out the following information:

* Commit id:
* Author:
* Date:
* Message:

Commit ids are usually very long, but you really only need the first 5 or so
characters to identify them. Verify this using `git show --name-only FULLID` and
`git show --name-only FIRSTFIVE`. Here "FULLID" means the entire commit id and
"FIRSTFIVE" is just the first five characters of that

The `git show` commands provides the details on the commit beyond just the log
info. Use `git show --name-only 14e2c4b` (you can also use your previous commit
if you already started this question).  What files where changed in this commit?

> YOUR ANSWER

Run `git show FIRSTFIVE`. What do you think this is telling you? What do you
think a line with a `+` means? What would a `-` mean?

> YOUR ANSWER


### Commiting to your local repository

By this point you have made some changes to this file. Save the file.

Go to the *Terminal* and type `git status`. What is this telling your?

> YOUR ANSWER

Use `git add` to place your changes in the staging area. Type `git status` again
to see what happened.

Suppose you decide your changes aren't a good idea. Use `git restore --staged`
to un-stage your changes. Verify the results again with `git status`.

Now add your homework back to the staging area and commit it with `git commit -m
"A COMMIT MESSAGE"`. What was the commit ID of your changes?

```

```

Demonstrate that you had a commit message using `git show --name-only HEAD` and
putting the results below. `HEAD` is shorthand for "the most recent commit".

```

```

If you forgot to include a commit message the first time, you can **amend**
your commit using
`git commit --amend -m "message"`. This can also be used to add files or changes
if you left them out of your staging area.

### Undoing commits

In the following box, write a sentence with a spelling or grammar mistake.

```

```

Use `git add` and `git commit` to commit this change. Include the commit ID:

```

```

Now use `git revert COMMITID`. What happens to this file?

> YOUR ANSWER

Correct the mistake and commit the changes. Demonstrate that the reverted change is no longer in the log using `git log -n 2` to show the last two commits.

```

```

### Working on branches and repositories

Before attempting this problem, make sure you have saved this file and committed
your changes (remember, you can check if there are any changes in your working
copy using `git status`).

If your terminal is currently in your repository, use `cd ..` to up by one
directory.

Now, we will pretend we are a coworker or collaborator getting access to your
repository.

Assuming you have checked out the repository to `stats306_fall_2022`, do the following
```
$ git clone stats306_fall_2022 stats306_copy
```

Use `ls stats306_copy` to see what is inside this new folder. What did you get?

> YOUR ANSWER

Save this file. Use `cd stats306_fall_2022` to go back into this directory and
commit the changes. Now use `cd ../stats306_copy` to go back to the other
repository. Use `git log` to see recent changes. Did the new changes show up
automatically? Use `git pull`. What do you see?

> YOUR ANSWER

We are now done with the copied repository. You may delete it if you wish using
`rm -rf stats306_copy`. Go back to your main repository.

You can also set up copies of your repository that you can share with others.
Git calls these copies "branches". Use `git branch -a` to see a list of branches
you currently have available. What branch are you own currently?

> YOUR ANSWER

Save your file and commit your changes. Create a new branch using

```
$ git checkout -b hw01
```

Use `git status` to verify that you are now on the `hw01` branch.

Make a file and commit it using:

```
$ echo "YOURNAME" > newfile.txt
$ git add newfile.txt
$ git commit -m "adding a file with my name"
```

(`echo` prints out something and the `>` directs it into a file)

Now, go back to the `main` branch using

```
$ git checkout main
```

Use `ls` and `git log` to verify that `newfile.txt` is not in this branch. You can even compare branches using

```
$ git diff HEAD hw01
```
Put the output here:

```

```

## Question 2 (5 points)

### Keeping code DRY (1 point)

A fundamental principle of good computer programming is avoiding repetition of
code. The acronym DRY stands for Don't Repeat Yourself. Clean up the code below
to avoid repetition.

```{r}
# A circle with diameter 7 has a circumfrence of
7 * pi

# A circle with diameter 7 has an area of
(7 / 2)^2 * pi

# A cylinder, with a face diameter of 7 and height of 16, has a volume of 
(7 / 2)^2 * pi * 16
```

### Mean and Variance (2 points)

The sample mean is defined as:

$$ \bar X_n = \frac{1}{n} \sum_{i = 1}^n X_i$$

and the sample variance is defined as
$$ \hat \sigma^2 = \frac{1}{n - 1} \sum_{i=1}^n (X_i - \bar X_n)^2$$

Using *vectorized* computations, compute the sample mean and sample variance of
the `displ` column of the `mpg` data set. Do not use the functions `mean` or
`var`. You may use `length` and `sum`.

### Starwars Characters (2 points)

Investigate the data set `starwars`. Answer the following

* How many characters are listed?
* What kind of data is the `mass` column?
* What kind of data is the `species` column?
* Explain what the value `NA` means in the `hair_color` column.
* How many characters are from Tatooine?
* For humans, what is the average (mean) age?

## Question 3 (5 points)

### Height and Mass (2 points)

Create a scatter plot of the columns `height` and `mass` from the `starwars`
data set. What jumps out from the plot? Do a little investigation to find out
about that point.

### Height and mass relationship (1 point)

Create a data set that excludes the outlier from the previous question. Plot the
relationship for height and mass again. Add a smoothed regression line.
Interpret these results for the nature of relationship between height and mass.

### Species (1 point)

The following code creates a new column in the `starwars` data that groups all
planets having only 1 character into an "Other" category.

```{r}
planet_counts <- table(starwars$homeworld)
starwars <- mutate(starwars, homeworld_other = ifelse(planet_counts[homeworld] == 1, "Other", homeworld))
```

Use this data set to make a bar plot of the `homeworld_other` column. Excluding
the "Other" category, what is the most common home world?

### More plotting (1 point)

Using the `starwars` data set, create a new plot that does not duplicate any of
the previous plots. Briefly describe the key elements of the plot and interpret
the results.

## Question 4 (5 points)

### Plotting housing prices (1 point)

Suppose we are interested in seasonal effects on the housing market. Here is a
plot of median housing prices by month in Texas over the period 2000 to
mid-2015.

```{r}
ggplot(txhousing, aes(x = month, y = median)) + geom_point() + geom_smooth()
```

What might be misleading about this plot? Refer to section 3.9 of R for Data
Science to propose a better way of displaying these data. Comment on any
seasonal effects you see.

### Prices by region (2 points)

Here is plot of median prices over time, attempting to breakdown the data by
region.

```{r fig.height = 10}
ggplot(txhousing, aes(x = date, y = median, color = city)) + 
  geom_point() + 
  theme(legend.position="bottom")
```

What would be a better way to display these data? Implement this plot. What area
saw the biggest increase in median price during this period? Which areas have
the largest variation in prices in a given year?

### Reducing volume of information (2 points)

Create a plot that summarizes the number of listings for each city. For each
city, show the minimum number of listings in a month, the maximum, and the
median.
