---
title: "Stats 306, Fall 2022, Homework 4"
author: "Your Name, Your Uniqname"
date: "Due November 7, 11:59PM"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

## Question 1

Let's look at counts of cancer deaths by type and several demographic categories
(age, race, sex).

```{r}
cancer_deaths <- read_csv("../data/AH_Provisional_Cancer_Death_Counts_by_Month_and_Year__2020-2021.csv.gz")
```

Due to the long names of the cancer categories, it is useful to the `kable` function to print out a nicely formatted table for several of these problems. Here is an example:

```{r}
select(cancer_deaths, 1:5) |> head(5) |> knitr::kable()
```

### Q 1.1

There are three groups of columns in this table:

- Date information
- Demographic information
- Cancer type deaths

There is also a redundant column that is the sum of the specific cancer type death rates. Remove this column from the table (and save the result to `cancer_deaths_2`). (Hint, you can use `- column` with `select` to remove a specific column)

After removing the redundant column, list the number date related columns, the specific demographic columns, and the number of cancer types.

### Q 1.2

Think about what is being measured here. Is this data in a tidy format (one row per observation, one column per measurement)? Use an appropriate `pivot_*` function to make this data more tidy (hint: see the help page for [tidy select](https://tidyr.tidyverse.org/reference/tidyr_tidy_select.html_ for some useful helper functions to pick all the cancer columns)). 

Call your table `cancer_tidy`. Use informative titles for the column names.

Display five rows of your table.

### Q 1.3

How many rows and columns are there in your `cancer_tidy` table? Multiply these values to get an approximate size of the data set.

For how many month-demographic-cancer combinations were there zero recorded deaths?

If we assume that any entry that is not included in the data set were zero, we could substantially shrink this data set. Filter the data set to exclude zero death entries. Call this `cancer_tidy_sparse`. How much smaller (as a percentage) is the sparse version?

### Q 1.4
Use the sparse version to answer the question, what were the top 5 forms of cancer by total deaths? 

## Q 1.5

Grouping by "Age Group" and cancer type, find the single most deadly cancer (by total deaths) in each group.  Do age groups vary in the most deadly form of cancer or all they all the same?


## Question 2

For this problem, we'll use the Lahman major league baseball package.

```{r}
if (!require(Lahman)) install.packages("Lahman", repos = "http://cran.us.r-project.org")
```

## Q 2.1

The `People` table in the Lahman package has information on individual players. There are two columns that are strings (characters) by default, but actually have date information: `debut` (date of first game played) and `finalGame`(date of last game played, or last game of 2021 regular season -- not including playoffs). 

Parse these dates into two new columns `debut_d` and `finalGame_d`.

Group the data by `finalGame_d` and count the nunber of players with each date as a final game. Sort in descending order. What do you notice about the date 2021-10-03? What is the maximum `finalGame_d` in the data set? How would you explain this result?

## Q 2.2

Create a new column in the `People` table called `career_length` which is the difference of `finalGame_d` and `debut_d`.

Show a density plot of this new column in years (using  `time_length(career_length, "years"))`. What is a typical career length?

# Q 2.3

Write a function that takes a date and returns the decade. For example, on the input `c("1891-07-10", "1997-10-01")` should return `c(1890, 1990)`. (Hint see the `round` function and its use of negative rounding.)

Compute the decade of each player's debut and plot a series of boxplots or violin plots of `career_length` by decade of debut. What trends do you notice? How do you explain the decrease in career length in recent years? (hint: what did you see in 2.1?)

# Q 2.4

Filtering out the players with `NA`s for `debut_d` or `finalGame_d`. Which player has had the longest career? (Display the name along with the duration in years.) Look up this player on Wikipedia to learn how he had such a long career. Was he playing the entire time? With which other player did he perform comedy routines?

For players continuing to play, we have a "censor" career length as we have not yet observed their total career, but such is the nature of time. Of players that have the maximum `finalGame_d`, which player has had the longest career? (Display the name and duration in years using the `time_length` function)


## Question 3

In this question we will continue to use the Lahman data.

## Q 3.1

Review the documentation for the `People` table and the `Fielding` table. Answer the following:

- What is the unit (row) for the `People` table?
- What is the unit (row) for the `Fielding` table? (Hint: see the code code below)
- What columns are common to both both tables?

```{r, cache = TRUE}
group_by(Fielding, playerID, yearID) |> summarize(years = n()) |> ungroup() |> summarize(sum(years > 1))
group_by(Fielding, playerID, yearID, stint) |> summarize(stints = n()) |> ungroup() |> summarize(sum(stints > 1))
group_by(Fielding, playerID, yearID, stint, POS) |> summarize(positions = n()) |> ungroup() |> summarize(sum(positions > 1))
```

## Q 3.2

Using a `inner_join`, find the names of the top 5 players for total double plays. (Hint: you will need to do some grouping and summarizing before your join.)

### Q 3.3

Review the Managers and AwardsManagers table.

Use a `left_join` to determine what proportion of managers won an award in each year. Graph this as a line plot. What do you notice about the proportion of managers winning awards?

Note: you will need to do some grouping first. What is this code snippet telling you:

```{r}
group_by(Managers, playerID, yearID, teamID) |> summarize(n()) |> nrow()
nrow(Managers)
```

## Question 4

### Q 4.1

Explain the difference between strings and factors in R. Explain why the following two sorts don't have the same result:

```{r}
v1 <- c("c", "b", "a")
v2 <- factor(c("c", "b", "a"), levels = c("c", "b", "a"))

sort(v1)
sort(v2)
```

### Q 4.2

Use the corpus of common words in `stringr::words`, create regular expressions that find all words that:

- Start with "y".
- End with "x"
- Are exactly three letters long. (Don’t cheat by using `str_length()`!)
- Have seven letters or more.

For each, provide the size of each set.

### Q 4.3

Continuing the the common words list, create regular expressions to find all words that:

- Start with a vowel.
- That only contain consonants. (Hint: thinking about matching "not"-vowels.)
- End with "ed", but not with "eed".
- End with "ing" or "ise".

### Q 4.4

Using a regular expression and `select`, find the columns in `cancer_deaths` that only refer to a single type of cancer.

