---
title: "Stats 306: Lecture 5"
subtitle: "Grouping, Summaries, Factions"
author: "Mark Fredrickson"
output: 
  slidy_presentation:
    incremental: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate) # install.packages("lubridate") if you don't have this
```

## Review: `ggplot`

* Using R's `factor` type to force categorical display (useful for color), some geometries
* Introduced formulas `y ~ x1 + x2` (models, delayed computation)
* Small multiples with `facet_wrap` and `facet_grid`
* Coordinate systems for display and certain data types.

## Review: operations on tables

* Tables: lists of vectors, `tibble` and `data.frame`
* Pulling out a single column: `d$col`
* Creating a single column: `d$newcol <- EXPR`
* Pulling out several columns: `select(d, col1, col2)` (and several other forms)
* Pulling out rows: `filter(d, criteria)`

## Temperature data for Ann Arbor, MI

```{r}
# to run this line from the console, use `setwd("lectures")` first
aatemp <- read_csv("../data/ann_arbor_weather.csv.gz")
```

## Plotting temperature vs. time

```{r}
ggplot(data = aatemp, aes(x = DATE, y = TMAX)) +
  geom_line()
```

## Days with large temp jumps

```{r}
temp_diff <- diff(aatemp$TMAX)
length(temp_diff)
dim(aatemp)
aatemp$temp_diff <- abs(c(NA, temp_diff))
```

## Distributon of differences
```{r}
ggplot(data = aatemp, aes(x = temp_diff)) + geom_bar()
```

## Plotting when large jumps occur during the year

Let's define a large jump in maximum temperature as being when one day is more than 10 degrees different than the subsequent day.

```{r}
ggplot(data = filter(aatemp, temp_diff > 10), aes(x = yday(DATE))) +
  geom_histogram() + coord_polar()
```

## Exercise

Combine `filter` and `select` to get only the columns `cty` and `hwy` for cars that have more than 4 cylinders.

```{r}
head(mpg, 1)
```

```{r filter-select, exercise = TRUE}

```
 
## Tasks
* Creating new column (in new table): `mutate(d, newcol = f(col1, col2))`
* Grouping and summarizing `group_by(d, discrete_column) |> summarize(new_name = a_function(col))`

## Mutate: create columns

Before we had code like:
```{r eval = FALSE}
df$new_column <- f(df$x, df$y)
```

It would be convenient to avoid the repeated `df$`:
```{r eval = FALSE}
new_df <- mutate(df, new_column = f(x, y))
```

## Mutate creates new tables

```{r}
dim(aatemp)
aatemp2 <- mutate(aatemp, tdiff = c(NA, diff(TMAX)))
dim(aatemp)
dim(aatemp2)
```

## Mutate for multiple columns

```{r}
aatemp3 <- mutate(aatemp,
                  tdiff = c(NA, diff(TMAX)),
                  tdiff_abs = abs(tdiff))
```

## Mutate to remove columns

```{r}
# NB: reassigning to same variable name
aatemp3 <- mutate(aatemp3, tdiff = NULL)
colnames(aatemp3)
```

## Using helper functions

Recall a $Z$-score is defined by:
$$Z = \frac{X - \bar X}{\hat \sigma}$$

```{r}
aatemp3 <- filter(aatemp3, !is.na(tdiff_abs))
aatemp3 <- mutate(aatemp3, z = (tdiff_abs - mean(tdiff_abs)) / sd(tdiff_abs))
ggplot(aatemp3, aes(x = z)) + geom_histogram()
```

## Conditional evaluation with `if_else`

We may want to create new values using a condition. The `if_else` and function can help:
```{r}
if_else(c(T, F, F), c("aT", "bT", "cT"), c("aF", "bF", "cF"))
```

R will also "recycle" values, so we can pass in single value that will get repeated:
```{r}
x <- c(-2, 1.4, -0.25, 7)
if_else(x < 0, 0, x)
```

There is also `ifelse`, which is similar, but a bit more permissive in what it allows for the two result vectors.

## Exercise

We often want to express variables on a different scale, such as constraining them to be between 0 and 1:
$$Y_i = \frac{X_i - \min(X)}{ \max(X) - \min(X)}$$

Use `mutate` to rescale `x` in this data: 
```{r rescale, exercise= TRUE}
d <- data.frame(x = rnorm(10))

```

## Exercise

Use `if_else` to replace any value greater than 1 with the value 1 and any value less than -1 with the value -1 (this is called "top coding").(*Hint*: you may want to do it two steps.)

```{r topcoding, exercise = TRUE}
d <- data.frame(x = c(-0.19, 1.35, 1.21, -0.11, -0.99, 
                      -0.4, -0.04, -0.4, 0.82, -1.55))

```

## `transmute`: `mutate` + `select`

If you only want the new column(s), you can use `transmute`:

```{r}
transmute(aatemp, degrees_from_freezing = abs(TMAX - 32)) |> summary()
```

## `summary` and `summarize`

R has a built in a function called `summary` that gives a distilled look at a table:
```{r}
aat_4col <- select(aatemp, c("STATION", "DATE", "TMAX", "SNOW"))
summary(aat_4col)
```

The `summarize` function is from `dplyr` (part of `tidyverse`) and allows computing arbitrary summaries.

```{r}
summarize(aat_4col, avg_TMAX = mean(TMAX), days_of_snow = sum(!is.na(SNOW)))
```

## Summarize variations: `_if`, `_at`, `_all`

```{r}
select(aatemp, where(is.numeric)) |> summarize_all(mean, na.rm = TRUE)
summarize_if(aatemp, is.numeric, var, na.rm = TRUE)
summarize_at(aatemp, c("TMAX", "TMIN"), c(maximum = max, minimum = min))
```

## Exercise

For the `mpg` data set, compute the mean `hwy` mileage and median `cty` mileage. Compute the variance of the ratio of `hwy` to `city`.
```{r summary, exercise = TRUE}

```

## Grouping

Often we want to break data out across categories and compute summaries within each.

```{r}
group_by(aatemp, year(DATE)) |> summarize(avg_TMAX = mean(TMAX), days_of_snow = sum(!is.na(SNOW)))
```

## Inspecting group data

```{r}

aat_year <- group_by(aatemp, year(DATE))
nrow(aat_year) == nrow(aatemp)
colnames(aat_year)[17]
group_vars(aat_year)
```

## Grouping by year and month
```{r}
aat_year_month <- group_by(aat_year, month(DATE), .add = TRUE)
group_vars(aat_year_month)
```

```{r}
aat_year_month <- group_by(aatemp, year(DATE), month(DATE))
group_vars(aat_year_month)
```

## Aggregating up with `summarize`

```{r}
summarize(aat_year_month, avg_TMAX = mean(TMAX)) |>
  ggplot(aes(x = `year(DATE)` + `month(DATE)` / 12, avg_TMAX)) +
  geom_line()

```

## Aggregating up two levels

```{r}
summarize(aat_year_month, monthly_avg_tmax = mean(TMAX)) |>
  summarize(yearly_median_monthy_mean = median(monthly_avg_tmax))
```

## Exercise

Using the `mpg` data set, find the manufacturer (`manufacturer`) with the highest mean highway efficiency (`hwy`)
```{r manufacturer-hwy, exercise = TRUE}

```

## Arranging output

Sometimes we want to choose the ordering of rows in a table. I don't use this a lot for raw data, but it can be quite helpful for summaries:

```{r}
group_by(aatemp, year(DATE)) |> 
  summarize(yearly_maxT = max(TMAX)) |> 
  arrange(yearly_maxT)
```

## Descending order, multiple columns


```{r}
group_by(aatemp, year(DATE)) |> 
  summarize(yearly_maxT = max(TMAX), yearly_minT = min(TMAX)) |> 
  arrange(desc(yearly_maxT), yearly_minT)
```

## Exercise

Group by both `manufacturer` and `class`. What manufacturer has the highest `cty` efficiency in the sense of the median of mean `cty` within class?

```{r manufacturer-hwy, exercise = TRUE}
```


## Useful functions for summaries

* Seen before: `mean`, `median`, `sd`, `min`, `max`
* Other common statistical measures: `quantile`, `IQR`
* For boolean/logical columns: `any` and `all` ("or" and "and" across vectors)
* The functions `n` and a `n_distinct` count units and distinct values

## Some more summaries

```{r}
summarize(aat_year, n(), n_distinct(TMAX), any(SNOW > 10))
```

## Centered temperature

```{r}
mutate(aatemp, TMAX_centered = TMAX - mean(TMAX)) |>
 ggplot(aes(y = TMAX_centered, x = factor(quarter(DATE)))) +
    geom_violin() 
```
## `mutate` and `group_by`

Observe some care when using `mutate` on grouped tables:

```{r}
group_by(aatemp, quarter(DATE)) |>
  mutate(TMAX_centered = TMAX - mean(TMAX)) |>
  ggplot(aes(y = TMAX_centered, x = factor(`quarter(DATE)`))) +
    geom_violin()
```
You may need to use `ungroup` to remove grouping information to get the behavior your want.

## Other R functions

Most other R functions do not recognize grouping:

```{r}
mean(aatemp$TMAX, na.rm = TRUE)
mean(aat_year$TMAX)
summarize(aat_year, mean(TMAX))
```

## Writing our own functions

```{r eval = F}
function_name <- function(required_argument, optional_argument = 0, ...) {
  # function body
  return(a_value)
}
```

(NB: The last line is automatically returned.)

## Exercise

Write an R function to transform a vector into a $Z$-score
```{r scale, exercise = TRUE}
x <- c(-1, 2, 1.1)
```

(NB: this function is called `scale` in R)

## Common use case: mutations and summaries

Suppose we want to median center and scale in terms of the IQR. We are going to use this functionality on the original data, but also within years.

```{r}
myscale <- function(x) {
  (x - median(x)) / IQR(x)
}
```
```{r}
mutate(aatemp, TMAX_myscale = myscale(TMAX)) |>
  ggplot(aes(x = factor(year(DATE)), y = TMAX_myscale)) +
    geom_violin()
```

```{r}
summarize(aat_year, myscale(TMAX)) |>
  ggplot(aes(x = factor(`year(DATE)`), y = `myscale(TMAX)`)) +
    geom_violin()
```

## Summarizing with functions that return more than one value

```{r}
mean_var <- function(x) {
  tibble(average = mean(x), variance = var(x))
}

summarize(aat_year, mean_var(TMAX))
```

Use caution when using this function more than once!

```{r}
summarize(aat_year, mean_var(TMAX), mean_var(TMIN))
```

## Predicates

A **predicate** is a function that returns `TRUE` or `FALSE`.

For example:
```{r}
no_missing <- function(x) {
  !(any(is.na(x)))
}

no_missing(c(2,3,44))
no_missing(c(7, NA, 122))
```

## Selecting columns with predicates

We have seen several predicates used already: `is.numeric` or `is.character` in selecting columns. We can also use our own predicates.

```{r}
select(aatemp, where(no_missing)) |> colnames()
```
## Anonymous functions

Sometimes we want to write short little functions that only get used once. We call these **anonymous functions** like this:

```{r}
select(aatemp, where(function(x) { !any(is.na(x)) })) |> ncol()
```

## Tidyverse anonymous function short hand

Many of the tidyverse functions support using formulas to indicate an anonymous function:

```{r}
select(aatemp, where(~ !any(is.na(.x)))) |> ncol()
```

>* `~` starts a formula
>* `.x` and `.y` are the arguments to the function

We'll see more of this form when we discuss iteration.

## Exercise

Let's put it all together using the `mpg` data set.

>* Get a list of manufacturers that produce cars in at least 2 different classes. (Recall `n_distinct` function)
>* Using that list, subset the `mpg` data to just those manufactures
>* Rescale the highway efficiency variable into Z-scores (using the common mean across all manufacturers)
>* Group the observations by manufacturer. Which one has the smallest variance in `cty` efficiency?

You may want to use `%in%`:
```{r}
c("Hi", "Low", "Low", "Medium") %in% c("Medium", "High")
```

```{r lastex, exercise = TRUE}

```

## Example solution

```{r}
at_least_2 <- group_by(mpg, manufacturer) |> 
  summarize(per_class = n_distinct(class)) |>
  filter(per_class > 1)

at_least_2

filter(mpg, manufacturer %in% at_least_2$manufacturer) |>
  mutate(cty_z = scale(cty)) |>
  group_by(manufacturer) |>
  summarize(v = var(cty_z)) |>
  arrange(v)
```

## Next time

>* R for Data Science: 19.4 - 19.6, 7 - 7.3
