---
title: "Stats 306: Lecture 12"
subtitle: "Loading Data, Dates"
author: "Mark Fredrickson"
output:
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate) # install.packages("lubridate") if you don't have this
library(nycflights13) # install.packages("nycflights13")
set.seed(2939394)
```


## Review

* Covariation
* Stratification and mixture distributions
* Lowess/smoothing
* Tables: matrix, tibbles, data frames

## Data Files

How do we share data?

* Plain text formats
* Proprietary formats
* Database applications

## Plain text delimited formats

* **Plain text** files are the easiest format to share across platforms, software.
* Somewhat less efficient for file size.
* We often **compress** the files using zip (`.zip`) or (`.gz`), but still basically plain text
* Organizing tabular data:
  * Each **line** of the file is a row of a table
  * Separate columns using a **delimiter**
  * Often need to **escape** strings if they contain delimiter
  
## Example, comma separated values (CSV)

```
-82.9,42.4,3043540,1703280231,Gratiot Ave & Mapleridge St,"ACCIDENT, HIT & RUN",OUIL,5422,54001,"ACCIDENT, HIT & RUN",2017/03/28 22:00:00+00
```

>* Columns separated by `,`
>* Numeric values interpreted as given
>* Both unescaped (`Gratiot Ave & Mapleridge St`) and escaped (`"ACCIDENT, HIT & RUN"`) strings.
>* Other kinds of data (e.g. date-times like `2017/03/28 22:00:00+00`) just represented as strings.

## Header rows

Most files will include a **header row** that gives column names, but not all!

Some files also have instructional text at the top which needs to be trimmed before loading.

## Loading delimited files

* R has a set of built in functions `read.EXTENSION` like `read.csv` and `read.table` (any white space as separator, "tab" is common).
* The built in functions return `data.frames`
* Tidyverse (in the )`readr` package) has `read_*` instead. Returns `tibbles` (has a few other features: faster, progress bars for big files, less data transformation initially)

## Example

(White space not important)
```{r}
ex_csv <-'
column1,column2,column3
7.1,sometext,01/02/2022
,"has,comma",07/15/2022
'
read_csv(ex_csv)
```

## Exercise

Use the `delim` argument to [`read_delim`](https://readr.tidyverse.org/reference/read_delim.html) to read this data into a table:
```{r doubledash, exercise = TRUE}
d <- '
col1--col2
7--8
--9
10--11
'
```

## Common issues

* Extra stuff at top of file, use `skip = NUMBER` to skip lines
* We assume column names by default, can indicate no names with `col_names = FALSE`
* Missing values: usually cells without values, but you can override with `na = "."`

## Investigating results

```{r}
ex_tib <- read_csv(ex_csv)
spec(ex_tib)
```

## Selecting column types

The `col_types` argument allows setting type of some or all columns. 
```{r}
read_csv(ex_csv, col_types = cols(column3 = col_date("%m/%d/%Y")))
```

See the help for the `cols` function for more. `col_skip` allows not importing column.

## Exercise

Read in the following table using `read_csv`. Make the first column be logical/boolean values. Make any "NONE" values be missing in R.

```{r settingcols, exercise = TRUE}
csv <- '
1,hello,NONE
0,goodbye,-3
1,NONE,1.111111
'
```
>* [`read_csv`](https://readr.tidyverse.org/reference/read_delim.html)
>* [`cols`](https://readr.tidyverse.org/reference/cols.html)
## Saving data

There are complementary `write_csv` and `write_delim` functions.

R has a proprietary format called "RDS". You can use `write_rds` to write a single table or the `save` function to write more than just one variable to a file. Useful for processing data in a file and then saving it for later use.

## Dates

One of the most common data-as-strings we need to deal with are **dates**, **times**, and **date-times**. 

There are lots of different formats, timezones, leapyears to deal with. The `lubridate` package helps:

```{r}
library(lubridate)
```

## Dates and date-times

```{r}
today()
now()
today() |> class()
now() |> class()
```

## The `mdy`, `ymd`, and `dmy` functions

If the dates are in one of the three most common conventions we have:
```{r}
mdy("10/01/2022") # Oct. 1
ymd("2022-10-01")
dmy("01.10.2022")
```

Can also read month names and abbreviations.

## Extensions to date-times

If we also have times, the `_hms` version will parse `HOURS:MINUTES:SECONDS` while `_hm` will just be `HOURS:MINTUES`. AM/PM can be included, otherwise 24 clock assumed. 

```{r}
mdy_hms("10/01/2022 12:01:37")
mdy_hm("10/01/2022 12:01am")
```

The UTC stands for "Coordinated Universal Time" (was Greenwich Mean Time)

We can select timezone using the nearest major city:
```{r}
mdy_hms("10/01/2022 12:01:37", tz = "America/Detroit")
mdy_hms("12/01/2022 12:01:37", tz = "America/Detroit")
```

## Exercise

Parse the following into dates or date times:
```{r parsedt, exercise = TRUE}
dt1 <- "31/10/89"
dt2 <- c("2022-07-01 3:10am",
         "1852-01-15 7:00pm")
```

## Parsing dates in files

We saw earlier the use of `cols(colname = col_date("EXPRESSION"))`. There is also `col_datetime` and `col_time`.

For these we need to give a pattern to match using various keys:

>* Year: four digit `"%Y"`, two digit `"%y"`
>* Month: number `"%m"`, abbreviation `%b`, full name `%B`
>* Day: number in month `"%d"`
>* Hours:minutes:seconds: `"%H:%M:%S"`
>* See `parse_datetime` for more

```{r}
parse_datetime("22/10/1, 10:01:00", 
               format = "%y/%m/%d, %H:%M:%S")
```

## Extracting components

```{r}
right_now <- now()
day(right_now) # same as mday()
month(right_now)
year(right_now)
yday(right_now) # day within year ("Julian date")
wday(right_now) # day within week, starting with Sunday
hour(right_now)
minute(right_now)
second(right_now)
```

## Time Math 

```{r}
ymd("2023/01/04") - mdy("12/16/2022")
```

The `seconds`, `minutes`, `hours`, `days`, `weeks` can be used to get offsets:
```{r}
right_now <- now()
right_now
right_now + minutes(10)
right_now + weeks(7)
```

## Durations

After doing math, you might want to convert the units.

```{r}
semester_end <- dmy_hms("16122022 23:59:59") - now()

(sed <- as.duration(semester_end))

as.numeric(sed, "days")
as.numeric(sed, "minutes")

```

There are equivalent `d*` functions (`dyears`, etc) for constructing durations directly.


## Difftimes vs. durations vs. periods

In the previous we did not really distinguish between three different classes of "time differences"

>* Difftime: R's built in method
>* Durations: differences in seconds
>* Periods: differences that are more human readble ("two months", which months?)

We'll leave further exploration as personal exploration.

## Exercise

Parse these dates.

If you wanted to set a reminder one month before each of these dates, when would the reminders be?

hat's the range of the dates? What is the mean date? What is the median date? 

```{r diffex, exercise = TRUE}
ds <- c("1981-09-25", "1982-03-06", "2010-12-14", "2012-10-18", "2019-04-03")
```

## Next time

R4DS: 12, 13


