---
title: "Stats 306: Lecture 6"
subtitle: "More Grouping and Summaries; Functions"
author: "Mark Fredrickson"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate) # install.packages("lubridate") if you don't have this
aatemp <- read_csv("data/ann_arbor_weather.csv.gz")
```

## Review: `mutate`

* `mutate`: Takes a table and returns a new table with columns added and/or subtracted
* Give us access to columns of `tbl` without having to write `tbl$column_name`
* `mutate(d, new_col = f(x, y), new_col2 = new_col1 + 1, old_col = NULL)`
* `if_else(cond, true, false)` (all are vectors of same length or single values): used `TRUE` and `FALSE` values in `cond` to pick value from `true` and `false` (respectively)
* Can use functions that summarize, e.g., `mutate(d, x_centered = x - mean(x))`

## Review: `groups and summaries`

* We often want to **partition** our data into smaller groups
* Things like finding **conditional means** or **conditional medians**
* `group_by` takes one or more grouping factors and creates a grouped table
* `summarize` computes group level summaries
* We'll see that `mutate` operates within groups, which can be quite helpful

## Temperature data for Ann Arbor, MI

```{r}
# to run this line from the console, use `setwd("lectures")` first
aatemp
```

## `mutate` to add a column

```{r}
aatemp_cel <- mutate(aatemp, TMAX_celsius = (TMAX - 32) * 5/9) |>
  select(TMAX, TMAX_celsius)
```

## Grouping by year
```{r}
aat_year <- group_by(aatemp, year(DATE))
summarize(aat_year, median(TMAX - TMIN, na.rm = TRUE))
```
## Useful functions for summaries

* Seen before: `mean`, `median`, `sd`, `min`, `max`
* Other common statistical measures: `quantile`, `IQR`
* For boolean/logical columns: `any` and `all` ("or" and "and" across vectors)
* The functions `n` and a `n_distinct` count units and distinct values

## Some more summaries

```{r}
summarize(aat_year, n(), n_distinct(TMAX), any(SNOW > 10))
```

## Centered temperature

```{r}
mutate(aatemp, TMAX_centered = TMAX - mean(TMAX)) |>
 ggplot(aes(y = TMAX_centered, x = factor(quarter(DATE)))) +
    geom_violin() 
```

## `mutate` and `group_by`

Observe some care when using `mutate` on grouped tables:

```{r}
group_by(aatemp, quarter(DATE)) |>
  mutate(TMAX_centered = TMAX - mean(TMAX)) |>
  ggplot(aes(y = TMAX_centered, x = factor(`quarter(DATE)`))) +
    geom_violin()
```

## Normalizing by monthly averages?

Let's center each observation by it's monthly average that we can understand if
it was unusual for that time of year.

```{r}
aat_month_centered <- group_by(aatemp, month(DATE)) |>
  mutate(TMAX_centered = TMAX - mean(TMAX, na.rm = TRUE)) # mean computed over months
## verify as the variance should be pr
summarize(aat_month_centered, var(TMAX_centered), sum(TMAX_centered^2) / (n() - 1)) |>
  head(3)
```

## Unusual months continued: conversion to ranks

*Ranks* are a useful robust replacement for values that are less susceptible to outliers. Let's rank days by how far they were from their monthly mean.

**Danger**: mutate will operate within months!

```{r}
mutate(aat_month_centered, r = rank(TMAX_centered)) |> 
  summarize(min(r), max(r))
```

## Ungrouping to fix

We need to drop the grouping values so that we can rank across all days.

```{r}
ungroup(aat_month_centered) |> 
  mutate(aat_month_centered, r = rank(TMAX_centered)) |>
  summarize(min(r), max(r))
```

## Average rank within years

Now that we can rank across all years and months, what year had the highest
average ranks?

```{r}
ungroup(aat_month_centered) |> 
  mutate(aat_month_centered, r = rank(TMAX_centered)) |>
  group_by(year(DATE)) |>
  summarize(mean(r)) |>
  arrange(desc(`mean(r)`))
```

## Exercise

Let's put it all together using the `mpg` data set.

>* Get a list of manufacturers that produce cars in at least 2 different classes. (Recall `n_distinct` function)
>* Using that list, subset the `mpg` data to just those manufactures
>* Rescale the highway efficiency variable into Z-scores (using the common mean across all manufacturers)
>* Group the observations by manufacturer. Which one has the smallest variance in `cty` efficiency?

You may want to use `%in%`:
```{r}
c("Hi", "Low", "Low", "Medium") %in% c("Medium", "High")
```

```{r lastex, exercise = TRUE}

```

## Example solution

```{r}
at_least_2 <- group_by(mpg, manufacturer) |> 
  summarize(per_class = n_distinct(class)) |>
  filter(per_class > 1)

at_least_2

filter(mpg, manufacturer %in% at_least_2$manufacturer) |>
  mutate(cty_z = scale(cty)) |>
  group_by(manufacturer) |>
  summarize(v = var(cty_z)) |>
  arrange(v)
```

## Other R functions

Most other R functions do not recognize grouping:

```{r}
mean(aatemp$TMAX, na.rm = TRUE)
mean(aat_year$TMAX)
summarize(aat_year, mean(TMAX))
```

## Writing our own functions

```{r eval = F}
function_name <- function(required_argument, optional_argument = 0, ...) {
  # function body
  return(a_value)
}
```

(NB: The last line is automatically returned.)

## Exercise

Write an R function to transform a vector into a $Z$-score
```{r scale, exercise = TRUE}
x <- c(-1, 2, 1.1)
```

(NB: this function is called `scale` in R)

## Common use case: mutations and summaries

Suppose we want to median center and scale in terms of the IQR. We are going to use this functionality on the original data, but also within years.

```{r}
myscale <- function(x) {
  (x - median(x)) / IQR(x)
}
```
```{r}
mutate(aatemp, TMAX_myscale = myscale(TMAX)) |>
  ggplot(aes(x = factor(year(DATE)), y = TMAX_myscale)) +
    geom_violin()
```

```{r}
summarize(aat_year, myscale(TMAX)) |>
  ggplot(aes(x = factor(`year(DATE)`), y = `myscale(TMAX)`)) +
    geom_violin()
```

## Exericse: Converting all Fahrenheit to Celsius

We have both `TMAX` and `TMIN` in Fahrenheit.

```{r}
select(aatemp, TMAX, TMIN)
```

Let's convert to Celsius ($C = (F - 32) \cdot 5/9$) by first creating a function.

```{r celsius, exercise = TRUE}

```

## Next time

>* R for Data Science: 19.4 - 19.6, 7 - 7.3
