---
subtitle: "Stats 306: Lecture 15"
title: "Relational Data, Continued; Introduction to Strings"
author: "Mark Fredrickson"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
set.seed(2939394)
if (!require(nycflights13)) install.packages("nyclfights13", repos = "http://cran.us.r-project.org")

sonnet18 <- "Shall I compare thee to a summer's day?\nThou art more lovely and more temperate:\nRough winds do shake the darling buds of May,\nAnd summer's lease hath all too short a date;\nSometime too hot the eye of heaven shines,\nAnd often is his gold complexion dimm'd;\nAnd every fair from fair sometime declines,\nBy chance or nature's changing course untrimm'd;\nBut thy eternal summer shall not fade,\nNor lose possession of that fair thou ow'st;\nNor shall death brag thou wander'st in his shade,\nWhen in eternal lines to time thou grow'st:\n\tSo long as men can breathe or eyes can see,\n\tSo long lives this, and this gives life to thee."

sonnet18_lines <- str_split(sonnet18, "\n")[[1]]
```

## Review

* `nycflights` data: flights, airlines, airports, planes, weather
* Keys: **primary keys** identify data within the table, **foreign keys** link to data in other tables.
* Relations: tables created by **joining** two or more tables. Classify as: mutating or filtering.
* Types of joins: outer: left, right, and fulll; inner
* Use of `dplyr` `*_join` functions, `by` argument, `suffix` argument

## Example: What models of planes fly from NYC to Detroit?

```{r}
filter(flights, dest == "DTW") |>
  left_join(planes, by = "tailnum") |>
  count(paste(manufacturer, model)) |> arrange(desc(n))
```


## Inner joins

Inner joins discard entries where a key does not exist in the table. Recall not all destinations are in the airports table (international flights):

```{r}
fiveam <- filter(flights, year == 2013, month == 1, day == 1, hour == 5)
left_join(fiveam, airports, by = c("dest" = "faa")) |>
  select(flight, dest, name)
```

Inner join would discard that row:
```{r}
inner_join(fiveam, airports, by = c("dest" = "faa")) |>
  select(flight, dest, name)
```

## Exercise

Use an inner join to find out how many domestic flights there were during the period covered by the `flights` data set.

```{r innerjoin, exercise = TRUE}

```

## Full joins

Left and right joins keep all rows in the respective table (discarding in the other freely). **Full joins** keep all:

```{r}
left <- head(flights, 10)
right <- filter(airports, faa %in% c("DTW", "BOS", "ORD", "LAX", "ATL"))
left_join(left, right, by = c("dest" = "faa")) |> select(flight, dest, name)
right_join(left, right, by = c("dest" = "faa")) |> select(flight, dest, name)
full_join(left, right, by = c("dest" = "faa")) |> select(flight, dest, name)
inner_join(left, right, by = c("dest" = "faa")) |> select(flight, dest, name)
```

## Multiple matches

Recall that we limited ourseleves to 5 airport codes. Notice the number of rows in this join:

```{r}
right_join(left, right, by = c("dest" = "faa")) |> select(flight, dest, name)
```

If A's key matches more than one row in B, A gets duplicated that many times. (This is different than `mutate`.)

Note, same behavior with a left join:

```{r}
# swapping order of table
left_join(right, left, by = c("faa" = "dest")) |> select(flight, faa, name)
```


## Keys with multiple columns

Examples thus far have used keys in single columns, but keys can be spread across columns (**compound**). 

```{r}
left_join(fiveam, weather, by = c("year", "month", "day", "hour", "origin")) |>
  select(flight, temp, wind_dir, wind_speed)
```

## Natural joins

To avoid lost of redundant typing, if two tables have a set of columns in common, those will automatically be used to join (**natural join**)

```{r}
left_join(fiveam, weather) |>
  select(flight, temp, wind_dir, wind_speed)
```

## Matching multiple keys

What would have happened if the `weather` table used the column name `airport` instead of `origin`? Since `airport` is not a column in `flights`:

```{r}
left_join(fiveam, weather, by = c("year", "month", "day", "hour")) |>
  select(flight, origin.x, temp, wind_speed, origin.y)
```

In a left join, if the same key matches multiple tables in the right, the row gets duplicated for each.

Lesson: Be careful with natural joins and compound keys.

## Filtering joins

A few times we used the fact that **outer joins** will result in tables with missing entries in order to count some feature.

This pattern is common enough to generate two **filtering join** functions:

- `semi_join(x, y)` keeps all observations in x that have a match in y.
- `anti_join(x, y)` drops all observations in x that have a match in y.

Big difference from mutating joins is that `y`'s columns are not included.

## Example

Suppose we want to find the top 10 destinations by number of flights:

```{r}
top_dest <- flights %>%
  count(dest, sort = TRUE) %>%
  head(10)
top_dest
```

If we *join* to the `top_dest` table, we can get flights to those destinations:

```{r}
left_join(flights, top_dest) |> # natural join on `dest` column
  select(flight, dest, n) -> flights_to_tops
flights_to_tops
```
Then filtering out the NAs
```{r}
filter(flights_to_tops, !is.na(n))
```

Do it one step with `semi_join`:

```{r}
semi_join(flights, top_dest) |>
  select(flight, dest)
```

## Exercise

Find the set of flights that were delayed by more than one hour (`dep_delay > 60`). Find the weather patterns associated with those flights and plot the distribution of wind speed with length of delay.

```{r semijoin, exercise = TRUE}
# you can use a natural join for flights and weather

```

## Other set operations functions

Base R has a series of **set operation** functions that operate on collects as mathematical sets (collections where items can appear at most once): `intersect`, `union`, `setdiff`, `setequal`, `is.element`, `%in%`

I frequently use these for doing comparisons of vectors:
```{r}
"DTW" %in% top_dest$dest # same as is.element
intersect(c("BOS", "LAX", "DTW"), top_dest$dest)
union(c("A", "B", "C"), c("Q", "A", "R"))
setdiff(c("A", "B", "C"), c("Q", "A", "R"))
setequal(c("A", "B", "C"), c("B", "C", "A"))
```

## Set operations with tables

With filtering joins, we can choose which columns link the tables. With the set operations, we must use all columns.

Flights to top destinations that are also delayed:
```{r}
flights_to_top <- semi_join(flights, top_dest) 
flights_delayed <- filter(flights, dep_delay > 0)
intersect(flights_to_top, flights_delayed)
```

## Relational data wrap up

* When data are stored in more than one table you need to **join**
* **Mutating joins** join tables to add columns (and some times rows!)
* **Filtering joins** use one table to select rows for another
* Outer left/right/full joins, inner joins
* Also useful for linking table to itself 
* Set operations less generally useful, but have some specific uses (e.g., vectors)

## Strings and Factors

R has two data types for representing categorical/nominal data types:

* Strings: character data
* Factors: integers representing categories with look up table for labels, can be ordered 

```{r}
example_str <- c("you", "say", "goodbye", "i", "say", "hello", "hello", "hello")
example_str
example_fct <- factor(example_str)
example_fct
```

## Strings vs. factors

Many times two types have the same function, why use one or the other?

- Want to order lexicographically: strings
- Want to order in other way: factors
- Treat as text (search, combine): strings
- Avoid errors from treating as text: factors

There used to be a space saving advantage to factors but that has been eliminated with newer versions of R.

## String basics

A string (or "character vector" in R) is character data enclosed in double `"` or single `'` quotes.

To include quotation marks inside, we need to **escape** them (like `,` in CSVs):
```{r}
baberuth <- "George Herman \"Babe\" Ruth Jr., a.k.a. \"the Bambino\" and \"the Sultan of Swat\""
cat(baberuth) # prints directly instead of displaying as escaped string
```

## Other special characters

Since `\` is used to escape things, if you need a literal backslash use `\\`. 
* New line: `\n`
* Tab: `\t`
* Unicode characters: `\unnnn` (u followed by 4 digit code)

This often come up in searching text for certain patterns.

## Combining strings

If we need to make strings out of other strings, we have two (more or less identical) functions:

* Old school: `paste` (joins with " ") and `paste0` (joins with ""); I frequently use these
* New school: `str_c` (joins with ""); used in our book

```{r}
paste("Section", c("100", "200", "300"), 
      "meets on", 
      c("Monday", "Tuesday", "Wednesday"))
```
Notice: this is a vector of length 3, use `collapse = "SEP"` to make a single string from a vector.

## Exercise

Join the following using a new line characters `\n` (use the `sep` and `collapse` arguments as necessary) and indent the last two lines using tabs (`\t`). Demonstrate useing `cat`.

```{r sonnet18, exercise = TRUE}
line1 <- "Nor shall death brag thou wander'st in his shade,"
line2 <- "When in eternal lines to time thou grow'st:"
line3 <- "So long as men can breathe or eyes can see,"
line4 <- "So long lives this, and this gives life to thee."
```

## Properties of strings

```{r, echo = FALSE}
line1 <- "Nor shall death brag thou wander'st in his shade,"
line2 <- "When in eternal lines to time thou grow'st:"
line3 <- "So long as men can breathe or eyes can see,"
line4 <- "So long lives this, and this gives life to thee."
```

The `length` function in R will tell you how many items are in a vector:

```{r}
sonnet18_end <- c(line1, line2, line3, line4)
length(sonnet18_end)
```

To find the length of the strings in terms of characters:
```{r}
str_length(sonnet18_end) # base::nchar also works
```

## Subsetting

The `str_sub` function can retrieve or change **substrings**.

```{r}
str_sub(line1, 11, 20)
str_sub(line2, 36)
str_sub(line2, -8)
line4b <- line4
str_sub(line4b, 1, 10) <- "REMOVED"
line4b
```

## Splitting

Here's the full sonnet as a single string:
```{r}
sonnet18 <- "Shall I compare thee to a summer's day?\nThou art more lovely and more temperate:\nRough winds do shake the darling buds of May,\nAnd summer's lease hath all too short a date;\nSometime too hot the eye of heaven shines,\nAnd often is his gold complexion dimm'd;\nAnd every fair from fair sometime declines,\nBy chance or nature's changing course untrimm'd;\nBut thy eternal summer shall not fade,\nNor lose possession of that fair thou ow'st;\nNor shall death brag thou wander'st in his shade,\nWhen in eternal lines to time thou grow'st:\n\tSo long as men can breathe or eyes can see,\n\tSo long lives this, and this gives life to thee."
```

Let's split that up into individual lines as strings in a vector:
```{r}
sonnet18_lines <- str_split(sonnet18, "\n") # notice use of new line
sonnet18_lines
```

What is happening? Since `sonnet18` could have had multiple entries (vector), `str_split` returns a **list** with splits for each item in the vector:
```{r}
str_split(c("a b c", "d e"), " ")
```

In this case:
```{r}
sonnet18_lines <- sonnet18_lines[[1]] # [[]] for lists
```

## Exercise

**Tokenization** is the process many computer languages use to turn strings into discrete words. Use `str_split` to turn each line of `sonnet18_lines` into individual words. Use `unlist` on the result and answer the question: how many words are there?

```{r tokenize, exercise = TRUE}

```


## Looking for patterns: Regexes

A **REG**ular **EX**pression (**regex**) is a mini-language for writing patterns to match in strings.

They are used in R but also in other languages, the command line, databases, etc.

A regex is just a string, albeit one that looks rather strange.

## Regexes pattern matches

Alphanumeric characters (a-z, A-Z, 0-9) match directly, most other characters with some later exceptions.

```{r}
str_view_all(sonnet18_lines, "thee")
```

`.` will match exactly character (not newline)

```{r}
str_view_all(sonnet18_lines, "th..")
```

Limited choices with `[]`
```{r}
str_view_all(sonnet18_lines, "th[aeiou][aeiou]")
```

Or classes of characters
```{r}
str_view_all(sonnet18_lines, "[A-Z]")
```

Negation choices with `[^]`
```{r}
str_view_all(sonnet18_lines, "th[aeiou][^aeiou ]")
```

## Other specials

To match periods we need to escape with `\\`

```{r}
str_view_all(sonnet18_lines[14], "\\.")
```

Beginning of string is "^", end of string is "$" (can get literals with escapes):

```{r}
str_view_all(sonnet18_lines, "^.. ")
```

Any white space (`\\s`) and make it optional `?`
```{r}
str_view_all(sonnet18_lines, "^\\s?.. ")
```

`\\d` matches digits.

## Repetition

* `?`: zero or one occurrence
* `*`: zero or many
* `+`: one or more

```{r}
str_view_all(sonnet18_lines, "m+")
```

## Groups and Alternation

To select between alternates use a vertical bar `|`. We often combine this with groups based on `()`:

```{r}
str_view_all(sonnet18_lines, "Thou|thou")
```

```{r}
str_view_all(sonnet18_lines, "fa(ir|de)")
```

Repeat patterns in `( )` using `\\NUMBER` (**backreferences**). Contiguous words with same letter: 
```{r}
str_view_all(sonnet18_lines, "(\\s|^)(.)[^ ]* \\2")
```
(notice "from fair" does not match because "fair from" already matched)

## Exercise

Find all contractions (words with "'" in them, but not possessive) in `sonnet18_lines` using `str_view_all`.

```{r contractions, exercise=TRUE}

```

## Next time

Using regular expressions for string parsing and other R tasks (14.4 - 14.6) and factors (15).
