---
subtitle: "Stats 306: Lecture 18"
title: "Iteration"
author: "Mark Fredrickson"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(nycflights13)
set.seed(2939394)

wikiart <- read_tsv("./data/WikiArt-Emotions-All.tsv.gz")
wikiart <- mutate(wikiart, Year = as.numeric(str_sub(Year, 1, 4)))

pivot_longer(wikiart, 
             matches(":"), 
             names_to = c("rating_type", "emotion"),
             names_sep = ":",
             values_to = "mean_annotation") -> wa_long

select(wa_long, Title, rating_type, emotion, mean_annotation)
wa_long <- mutate(wa_long,
                  emotion = str_replace(emotion, "\\s", ""))

filter(wa_long, str_detect(rating_type, "Art")) |>
  group_by(ID) |>
  arrange(desc(mean_annotation)) |>
  summarize(strongest_emotion = first(emotion),
            strongest_emotion_value = first(mean_annotation)) ->
  wa_art_strongest

wa_art_strong_year <- left_join(wikiart, wa_art_strongest, by = "ID")
```

## Review

* Wikiart data: made tidier and cleaner using **regular expression** (regexes)
** `seclect` and `pivot_longer` with `matches`
** `str_replace`
* Using tidied data for summaries, plots
* Grouping data by art to find strongest emotion (high mean annotation)
* `left_join` to get year and categorization information

## Factors

* Collections of labels, fixed set of possibilities
* Can set order of labels and change labels easily
* `factor(vector_of_strings, levels = optional_ordering_of_levels)`
* reordering existing factors with `fct_reodrer` or `fct_infreq`
* merge with `fct_collapse`

## Solving Repetition Issues

We used functions to reduce repetition in code since the beginning of class.

```{r}
x <- c(10, 1, 23)
xbar <- mean(x)
sdx <- sd(x)

## bad!
c((x[1] - xbar)/sdx, (x[2] - xbar)/sdx, (x[3] - xbar)/sdx)

## better
f <- function(x) { (x - xbar) / sdx }
c(f(x[1]), f(x[2]), f(x[3]))

## best
g <- function(x) { (x - mean(x)) / sd(x) }
g(x)
```

## Vectorization and Its Limits

The final solution used **vectorization**: repeating the same calculation for every item in a vector.

Most R functions are vectorized, but every now and again, we encounter results that can be easily vectorized.

* Computing the same calculation over each column in a table (ignore `mutate_all` for the moment)
* Compute values over all possible combinations of two vectors
* Build up a result by combining the values one at a time other calculations where order is important

## Unvectorizable computation

An **outer product** of two vectors (`x` with $n$ elemnts and `y` with $k$ elements) is the $n\times k$ matrix with entries $M_{ij} = x_i \cdot y_j$.

So if we had
```{r}
x <- c(1, 2, 3)
y <- c(10, 20, 30, 40)
```

We could create the outer product using:
```{r}
matrix(c(x[1] * y, x[2] * y, x[3] * y), ncol = 4, byrow = T)
```

## `for` loops

```{r, eval = FALSE}
for (variable_name in collection) {
  ... code that uses variable_name ...
}
```

* The `seq_along` function can be used to get indexes in to a collection (like items in a vector or columns in a table)
* If we need to save results, we often create the data first using `vector("type", integer_length)`
* We could also use `results <- c(results, new_value)`, but this is a less efficient method

## Outer product with for-loop

```{r}
## version 1
result <- matrix(0, nrow = length(x), ncol = length(y))
for (i in seq_along(x)) {
  for (j in seq_along(y)) {
    result[i, j] <- x[i] * y[j]
  }
}
result

## version 2
result <- matrix(0, nrow = length(x), ncol = length(y))
for (j in seq_along(y)) {
  result[, j] <- x * y[j]
}
result

## version 3
result <- numeric(0)
for (y_j in y) {
  result <- c(result, x * y_j)
}
result
matrix(result, ncol = length(y))
```

## Exercise

Using the function `class`, determine the type of each column in `nycflights13::flights` using a `for` loop. (Recall, tables are lists of vectors, so you can get a column out using `tbl[["colname"]])`)

```{r flightsclass, exercise = TRUE}
class(nycflights13::flights$tailnum)
```

## Exercise

Write your own version of the `min` function using a `for` loop.

```{r minfun, exercise = TRUE}

``` 

## `while` loops

In some cases we don't know the length of the loop in advance and we need to check to see if we should stop.

Find the first instance of an odd number in the vector:

```{r}
x <- c(2 * 1:5, 11, 13, 4)
x
pos <- 1
while (x[pos] %% 2 == 0) {
  pos <- pos + 1 # pos++ like C/C++
}
pos
x[pos]
```

## `map` functions

Many loops have the format
```{r eval=F}
result <- vector("type", length(x))
for (.x in x) {
  result[i] <- ... something with .x ...
}
```

We can avoid this repetition by using `map` functions to handle setting up the storage, running the loop, saving the results so we can write

```{r eval=F}
map(x, ~ ... something with .x ...)
```

## Compound Random Variables

Many times we take sample sizes $n$ to be fixed. But what if our sample size itself was random? 

A **compound random variable** is one in which a discrete random variable sets the sample size, and then some other random variable is replicated that many times.

```{r}
r_pois_norm <- function(n, lambda = 1, mu = 0 , sd = 1) {
  replicate(n, {
    n_i <- rpois(1, lambda) + 1
    rnorm(n_i, mu, sd)
  })
}

samples <- r_pois_norm(3, lambda = 8)
samples
```


## Maximum of the samples

What was the largest value of each sample?

```{r}
map(samples, max)
```

What was the largest magnitude?
```{r}
## version 1
map(samples, ~ max(abs(.x)))
```
Recall: `~ ... operations on .x ...` is equivalent to 
```{r eval = F}
function(.x) { ... .x ...}
```

```{r}
## version 2
map(samples, abs) |> map(max)
```

## Exercise

Use `map` to get the classes of each column of the `nycflights13::flights` table.

```{r mapclass, exercise = TRUE}

```

## Exercise

Write a function that takes a column as input and either

* Returns a short summary in the form of the mean, median, and standard deviation if it is `numeric`.
* Returns the class of the object otherwise

Use map to apply it to all columns of the `nycflights13::flights` table.
```{r mapsummary, exercise = TRUE}
is.numeric((c(0.1, 20)))
```

## Getting vector outputs

`map` can take either a list or a `vector` as input, but always returns a `list` as output. If you know that all outputs will be the same, you can `unlist` to get the values in the form you want:

```{r}
map(samples, ~ max(abs(.x))) |> unlist()
```

## Known types

If you know that all of the entries in the list will be of a particular type, you can `flatten_*` them to that type:

- `dbl`: Numeric data, decimal (can handle integers too)
- `int`: Numeric data, integers
- `lgl`: Logical/boolean
- `chr`: Character/strings 

```{r}
map(samples, ~ max(abs(.x))) |> flatten_dbl()
```

You get an error if things don't match:
```{r, eval = FALSE}
flatten_dbl(list(c("hello")))
# Error: Can't coerce element 1 from a character to a double
```

## `map_*` versions

The pattern `map(input, f) |> flatten_*()` is so common, there are parallel versions for map:

```{r}
map_dbl(samples, ~ max(abs(.x)))
```

## Exercise

Use a `map_*` version (`lgl`, `chr`, `dbl`, `int`) to get classes of the columns in `nycflights12::flights`. 

```{r mapunderscoreclass, exercise = TRUE}

```

## Additional arguments

We have seen several times that R functions take `...` and can use that to pass additional arguments along  to other functions.

```{r}
samples_missing <- samples
samples_missing[[1]][1] <- NA

map_dbl(samples_missing, mean)
map_dbl(samples_missing, mean, na.rm = TRUE)
```

## Mapping on two things

Suppose for each sample, we have another outcome:

```{r}
samples_y <- map(samples, ~ runif(length(.x)))
samples_y
```

What is the difference corrleation within each sample?

```{r}
map2(samples, samples_y, cor)
map2_dbl(samples, samples_y, cor)
```


## Mapping your own function on two things

Suppose we wanted the average ratio of X to Y within each sample. We need to write a custom function with two arguments:

```{r}
map2_dbl(samples, samples_y, function(x, y) { mean(x / y)})
map2_dbl(samples, samples_y, ~ mean(.x / .y))
```

## Exercise

Using `map2`, get the first word of each item in `stringr::fruit` (keep single words). 

Some useful examples:
```{r}
str_locate("chili pepper", " ")
str_sub("chili pepper", end = 6 - 1)
str_locate("canataloupe", " ")
```


```{r firstword, exercise = TRUE}
first_space <- str_locate(stringr::fruit, " ")[,1]
```


## Other loop replacing functions

`keep` takes only the entries that match a predicate:

```{r}
keep(samples, ~ length(.x) > 10)
```

`discard` drops entries that match.

Many `for` loops have **accumulators**:

```{r}
max_length <- -Inf
for (i in seq_along(samples)) {
  max_length <- max(length(samples[[i]]), max_length)
}
max_length
```

`reduce` avoids the bookkeeping (often convenient to map first):
```{r}
map_dbl(samples, length) |> reduce(max)
```

We can see the intermediate results with `accumulate`:
```{r}
reduce(c(1, 10, 100, 2, 3, 4), `+`)
```

## Map functions that return tables

The `_dfr` variants use functions that return data frames, and then row bind them together
```{r}
map_dfr(samples, ~ tibble(x = .x))
```
We probably want to know which sample each value is so we can do operations after. We can use an **indexed map** with the `imap_*` variants:
```{r}
imap_dfr(samples, ~ tibble(x = .x, sample_num = .y)) |> 
  group_by(sample_num) |> 
  mutate(x_centered = x - mean(x)) # within group, mean centered values
```

There is also `_dfc` that binds columns together.


## Connections to grouping and summary functions

Many of these tasks can also be done by creating tables and running `group_by` and `summarize`. There is also a `group_map`

```{r}
group_by(wikiart, Category) |> group_map(function(x, grp) { 
    if (nrow(x) > 10) {
      return(mean(x$`Ave. art rating`))
    } else {
      return(max(x$`Ave. art rating`))
    }
  }) |> flatten_dbl()
```

## Built in R functions

There are several built in functions that are similar to `map_*`:

* `lapply` is similar to `map`, but doesn't allow for the function shorthand notation
* `sapply` is similar to the `map_*` varieties, but no checking. Will return a list if it can't safely unlist results.
* `vapply` is even closer to `map_*`, but still no anonymous functions
* `mapply` is similar to `map2`

R also has some capitalized versions (`Map`, `Reduce`), but generally no checking less convenient to use with pipes.

## Errors in iteration

If we run into a bug in the code being iterated, it can often be difficult to find the source.

```{r}
x <- list(1, 10, "3", 7)
```

```{r eval = FALSE}
x <- list(1, 10, "3", 7)
map(x, log)
# Error in .Primitive("log")(x, base) : 
#  non-numeric argument to mathematical function
```

The `safely` function catches these errors and gives back useful info:

```{r}
safe_log <- safely(log) # returns a new function
safe_log(10)
safe_log("3")
map(x, safe_log)
```


## `transpose` function

Recall our model of a table:  a list of length n composed of vectors of length k. If we wanted to **transpose** the table we would have a list of length k with vectors of length n.

The `transpose` function does this for data that can be thought of as a table:
```{r}
x_log_tp <- transpose(map(x, safe_log))
x_log_tp$result
x_log_tp$result |> flatten_dbl()
x_log_tp$error
```







