---
subtitle: "Stats 306: Lecture 23"
title: "Shiny: Data dashboards in R"
author: "Mark Fredrickson"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate)
library(shiny)
```

## Review

* Parallel programming: split up work for multiple cores/CPUs/nodes
* `makeCluster`, `clusterExport`,  `clusterEvalQ`
* `parLapply`, `parSapply` (works on Windows, clusters of multiple computers)
* `mclapply` (Mac/Linux shared memory)
* Rcpp: Simplified C/C++ interface for R
* Great for loops, recursive functions, costly R execution

## Shiny

We've used RMarkdown to create **static** displays of data. To make **interactive** displays, we'll use the **Shiny** package.

We break up Shiny apps into three pieces:

* User Interface (UI): handles the visual presentation and control interface
* Server: Processes input and creates data for display
* Reactive programming: how things change when users modify app

NB: We've been using Shiny for our interactive exercises in slides.

## Web app basics

The web is based on some basic technologies that make interoperation possible:

* Web server to handle **requests** (server side)
* Web browser to display web pages (user side)
* Web pages written in HTML (hypertext markup language), e.g., `<p>This is a paragraph with <b>bold</b> text and a <a href="URL">link</a>.</p>`
* Cascading style sheets (CSS) to add layout and display directives
* User side vs. server side code
  * User side: JavaScript executes within the browser
  * Server side: Many possible languages (PHP, Python, JavaScript/node.js, .NET, many others )
* Databases, images, etc.

## Creating an app

RStudio: `File -> New Project`

## Running Apps

* Starts a local web server (typically private with most PC firewalls)
* Can view in R or a web browser by going to `127.0.0.1:XYZS`
* Can deploy to public web servers

## User Inteface

The heart of the Shiny app is the `shinyApp(ui, server)` function. Let's start with the `ui` argument.


Goal: define the set of pieces the user will interact with and observe.

Shiny generally tries to give reasonable default layouts that will resize with different user screens/environments.

## Pages

Shiny can work with **single page** applications or **multipage** applications. A **page** is the canvas of the app.

* `fluidPage`: Usually the best option for setting up a page, allows for components to fill the screen horizontally
* `fixedPage`: If you want to limit the maximum (horizontal) size of the page.
* `fillPage`: fills the screen vertically

The `fluidPage` page takes a `...` argument which will constitute the set of **panels** from top to bottom of the page.

## Sidebar layout

A common layout for Shiny apps has a title at the top, a side panel on the left, and a main panel on the right.

```{r eval = FALSE}
fluidPage(
  titlePanel(
    # app title/description
  ),
  sidebarLayout(
    sidebarPanel(
      # inputs
    ),
    mainPanel(
      # outputs
    )
  )
)
```

See `clt_app.R` for a simple example.

## Creating your own layouts

`sidebarLayout` is a convenient wrapper around the `fluidRow` function. Each **row** is composed of 12 **columns**. 

For example
```{r eval = FALSE}
fluidPage(
  fluidRow(
    column(6, ...)
    column(6, ...)
  ),
  fluidRow(
    column(10, ...)
    column(2, ...)
  )
)
```

See `clt_lln.R`

## Interactive Graphics

Shiny tracks clicks across the page and can provide information on where within a graphic the user clicked.

```{r eval = FALSE}
plotOutput("plot", click = "plot_click") # populates input$plot_click
```

## `req` function

A useful function to have for only including server logic when input is present is `req`. 

If the data does not exist, `req` terminates execution of the rest of the code in that `render` function.

See `click_graph.R`


## Brushes and `tableOutput`

We can draw rectangles on the graphic using the `brush = ` option to `plotOutput`.

The `brushedPoints` function takes a brush and a table returns the points inside of the rectangle.

See `brush_graph.R`

## Exercise

Cut the mean mean and variance of the points included in a brush.

See `brush_example.R`.

## User input

Basic HTML includes a few basic building blocks for user input:

* Text boxes (single line, multiline)
* Check boxes
* Radio buttons (a set where (at most) one can be selected)
* Drop down menus
* File uploads
* Clickable buttons ("Submit" or "Cancel")
* Hidden fields

Shiny adds some other useful features

* Sliders
* Date/time input
* Parsing numbers

Lets go look at the [Shiny Widget Gallery](https://shiny.rstudio.com/gallery/widget-gallery.html)

## Basic inputs

Boxes for inputs allow users to type in values.

Both `textInput` and `numericInput` take three arguments:

* `inputId`: the entry in the `input` list used by `server`
* `label`: The text that the user will see prompting input.
* `value`: A default value.

`textInput` can also have `placeholder` text inside. `numericInput` can have some basic limits (min, max, step size)

See input_song.R

## Exercise

Open `example_zoom.R`. Add four inputs for numeric values to get the bounds of the plot. Then update the ploting code to use those bounds. 

