---
title: "Stats 306: Lecture 1"
subtitle: "Basic R and RMarkdown"
author: "Mark Fredrickson"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
```

## Review 

* Course overview, objectives, policies
* Software: R, RStudio, Git
* Great Lakes cluster environment
* Command line interface (CLI)

## Git and the CLI

When interacting with git, we will use commands like this:

```
$ git SUBCOMMAND -o --options ARG1 ARG2 ...
```

Some example subcommands will be `pull` to get updates, `log` to see the history, `add` to stage files, and `commit` to commit changes.

## Let git know who you are

Last time we saw

* `git config`
* `git clone`

## Getting future updates

```
$ cd PATH/TO/REPO
$ git pull
```

## What can we do with the repo?

>* Run these slides
>* Edit an example homework and save changes to your local repository.

## RMarkdown documents

RMarkdown documents blend **code** and **presentation**. They will be the tool we use for

>* Lecture slides
>* Labs
>* Homework assignments

HW Workflow: edit document, knit, submit HTML 

## What is markdown?

* Markdown files are **plain text** files that have special interpretation.
* Came out of common conventions for adding expression to emails.
* They have a header (denoted by `---`) and a body. We can **render** them to HTML to view in any web browser.
* See the "Help -> Markdown Quick Ref" in RStudio for more details.

## RMarkdown

* RMarkdown extends basic markdown to allow embedding R code within the document.
* It addition to formatting text, you can include code **chunks**.
* When you **knit** the document, first the code is evaluated, then the results are added to the markdown document. Then that is turned into a web page.
* In addition to text and numeric output, plots and graphs are automatically inserted into the output.
* There are options to control how chunks are presented, cache computations to speed up knitting, etc.

## Chunks in Rmd files

All chunks have the following form:

<!-- The extra stuff is to get this print out as if
it were an R chunk, even though it is not really one. -->
<pre><code>```{r name, key = value, ...}
... R code goes here ...
<span>```</span></code></pre>

* The `name` is optional but helps when you get an error.
* The `key = value` controls the behavior of the chunk (optional). Example, `include = FALSE` to hide chunk.
* `CTRL-ALT-I` inserts a new chunk your RMarkdown document.

## Working on the example HW again

Task: Let's answer a few questions and prepare our HW for submission.

## Saving changes to our local repo

```
$ git status
$ git add FILENAME
$ ... perhaps other files ...
$ git commit -m "a note explaining changes"
```

## R's Read-Evaluate-Print-Loop (REPL)

* The **Console** in R lets you interact with a running R process.
* When you type in in the console, R **reads** the command to make sure it is valid R code.
* R **evaluates** the valid code -- produce the result of the computation
* R **prints** the result back to the console.
* The process then repeats (**loop**)

## Trying out the REPL

We will use exercises in the slides like a REPL. Use the math operations `*`, `\`, `+`, and `-` to compute twice the difference of 100 and 18. You can use `()` to group calculations.

```{r repl, exercise = TRUE, exercise.lines = 1}

```
The "Start Over" button will give you a fresh environment.

## Expressions and Statements

If you stopped a running R program at some point in time, everything about the program is called its **state**.

An **expression** is R code that, when run, provides a value (do not change state):

```{r eval = F}
3 + 4
```

Commands that change the **state** of the program are called **statements**, such as assignment:

```{r}
a <- 3 + 4
```

## Variables and Assignment

A **variable** is a box to store data. We put the results of expressions into variables using **assignment**.
```{r}
a <- 7
b <- 12
c <- a + b
print(c)
```

Rules for variable names: start with a letter then letters, numbers, `.`, `_`. 

## REPL and State

Suppose you entered the following at your R console:
```{r}
a <- 3 + 1
a <- 10 + 2
```
```{r repl-state-setup}
a <- 3 + 1
a <- 10 + 2
```
What would happen if you did the following?
```{r repl-state, exercise = TRUE}
print(a)
```

## The REPL and Chunks

When you **knit** an RMarkdown document, RStudio **starts a new process**.

Every time it encounters a **R chunk** it **evaluates** the R code inside and inserts the output into the document (default; can be changed). Like a behind the scenes REPL.

After running all code, it turns the markdown into a web page (`example-homework.Rmd` => `example-homework.html`)

## Example of chunks and state

(Feel free to look at `lecture01.Rmd` right now).

Here is a code chunk:
```{r}
classname <- "Stats 306"
```
Here is some intervening text and a second chunk:
```{r}
print(classname)
```

## Order is important

Because **statements** change the context, order of statements is very important. Compare
```{r}
a <- 7
print(a)
a <- 2
```

```{r}
a <- 7
a <- 2
print(a)
```

## Working with code chunks, debugging hint

You can send a line of code to the console using `CTRL-Enter`.

You can run all of the code in a chunk at once using the green arrow or `CTRL-SHIFT-Enter`.

If you run into a problem, use the `Session -> Restart R` to clean up, and the use "Run all chunks above" (`CTRL-ALT-SHIFT-p`). Then try the chunk again.

## What can go in variables?

R has several different **data types**. Some common types:

* **Numeric**: `7`, `3.22`, `9.1e2`
* **Strings**: `"Hello"` or `'Hello'` (also called "character vectors")
* **Logicals**: `TRUE` and `FALSE`
* We'll see more data types going forward
* These are also referred to as the **class** of the data.

## Why are variables useful?

* Avoid repeatedly typing in the same thing
* Save the results of computations
* Avoid recomputing same values
* Separate what we are computing from any particular input (**abstraction**)

## Exercise: Using variables

Find the difference between 193 and 7. What is this value divided by two? What is this value divided by three? (Only compute the difference once.)
```{r using-variables, exercise = TRUE, exercise.lines = 5}

```

## Style point: descriptive variable names

* So far we have used names like `a` and `b`.
* For large programs, having variable names that describe what they hold is much better
* Compare `a` to `number_of_seconds_in_a_day`
* In RStudio, it's just as easy to type with `TAB` completion
* On HW assignments, GSIs will be awarding points for *style* (readable code)

## Using variables: functions

Once we have data stored in a variable, one of the main ways to use it is to **call a function.**

```{r, eval = FALSE}
function_name(argument1, argument2, named_argument = additional_data, ...)
```

## What do functions do?

* Functions behave (mostly) like mathematical functions $f(x)$.
* They take **arguments** and return **values**
* Encapsulate code (mini-programs)
* Make it easy to repeat operations

## Converting from degrees to radians

Recall two ways to measure an angle:

>* Degrees: 0 to 360
>* Radians: 0 to 2$\pi$

Exercise: compute the radian equivalent of 45 degrees (use a good variable name!). Then use the `sin(your_variable_name)` function to compute the sine of 45 degrees. (Hint: there is a built in variable named `pi`.)

```{r sin45, exercise = TRUE}

```

## Math operations are functions too

Suppose we had `mult(x, y)` and `div(x, y)`.

We could have written the previous calculation as 

```{r eval = FALSE}
mult(2, mult(pi, div(45, 360)))
```
Those aren't real functions, but R transforms `2 * pi * 45/360` into that behind the scenes.

We call functions that do not have explicit function calls **operators**. See help pages for `Arithmetic Operators`, `Relational Operators`, and `Logical Operators` for some of the most common.

## Function evaluation

In R, when we call a function, first the **arguments** are evaluated, then the function is called.

```{r eval = FALSE}
sin(mult(2, mult(pi, div(45, 360))))
```

* Compute `value1 <- div(45, 360)` 
* Compute `value2 <- mult(pi, value1)`
* Compute `value3 <- mult(2, value2)`
* Compute `sin(value3)`

## Side effects

Some functions also have **side effects** beyond just returning a value.

* Load or save a file
* Take input from a mouse
* Display something on the screen

## The `curve` function

The `curve` function is a useful way to graph out a function:
```{r, height = 4, width = 7}
curve(sin(x))
```

Here `x` is a "special variable" that does not exist outside the use of `curve`.

## Changing the bounds of the plot

`curve` can take more than just one argument,

```{r, height = 3, width = 5}
curve(sin(x), 0, 2 * pi)
```

## Optional and named arguments

R allows you to use the name of arguments to pass them in any order:

```{r}
curve(to = 2 * pi, expr = sin(x), from= 0)
```

By *default* `from = 0` and `to = 1`


## Avoiding making variables we don't need

Sometimes we only need to use something once, so taking the time to come up with a variable name is unncessary. For example,

```{r}
radians_45_degrees <- 2 * pi * 45/360
sin(radians_45_degrees)
```

Instead we can use **pipes** to pass the first argument of function

```{r}
(2 * pi * 45/360) |> sin()
```

(NB: See the help page for `Syntax` for operator precedence.)

## R's assignment operators

R has two ways to store data in variables:

```{r}
variable_arrow <- 7
variable_equal = 10
variable_arrow + variable_equal
```

I like the arrow because it can be reversed
```{r}
7 -> variable_reversed
pi * variable_reversed
```

## Some special values

In addition to `pi`, R has some other built in variables (see help page for `Constants`).

Also several special data values

* `NA` means "missing". Including a `NA` value in a calculation will usually make the result also `NA`.
* `NULL` means "not set yet" and is often used for optional arguments to functions
* `NaN` means something was divided by zero or other mathematical impossibility.

What will happen with this code?

```{r na-prop, exercise = TRUE}
angle_degrees <- NA
angle_radians <- 2 * pi * angle_degrees / 360
sin(angle_radians)
```

## Resassigning values

What happens will the following code produce?
```{r reassign, exercise = TRUE}
some_data <- "SECRET CODE"
some_data <- 777
print(some_data)
```

## Summary

* Git and the command line
* RMarkdown documents: text and code chunks
* R and the REPL
* Variables: boxes to hold data
* Functions: little programs that operate on data

Next time: Lab on Tuesday, we will discuss more about R data next Wednesday
