---
subtitle: "Stats 306: Lecture 13"
title: "Tidy Data"
author: "Mark Fredrickson"
output:
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
set.seed(2939394)

table1 <- bind_cols(tidyr::table1, continent = c("Asia", "Asia", "South America", "South America", "Asia", "Asia"))

table2 <- pivot_longer(table1, c("cases", "population"))

mi_population <- structure(list(name = c("Detroit", "Grand Rapids", "Warren", 
"Flint", "Sterling Heights", "Lansing", "Ann Arbor", "Livonia", 
"Dearborn"), county = c("Wayne", "Kent", "Macomb", "Genesee", 
"Macomb", "Ingham, Clinton, Eaton", "Washtenaw", "Wayne", "Wayne"
), `2010` = c(713777, 188040, 134056, 102434, 129699, 114297, 
113934, 96942, 98153), `2000` = c(951270, 197800, 138247, 124943, 
124471, 119128, 114024, 100545, 97775), area_km2 = c(359.4, 115, 
89, 86.6, 95, 93.4, 72.1, 92, 62.7)), row.names = c(NA, -9L), spec = structure(list(
    cols = list(name = structure(list(), class = c("collector_character", 
    "collector")), county = structure(list(), class = c("collector_character", 
    "collector")), population...3 = structure(list(), class = c("collector_number", 
    "collector")), population...4 = structure(list(), class = c("collector_number", 
    "collector")), area_km2 = structure(list(), class = c("collector_double", 
    "collector"))), default = structure(list(), class = c("collector_guess", 
    "collector")), delim = ","), class = "col_spec"), problems = NULL, class = c("spec_tbl_df", 
"tbl_df", "tbl", "data.frame"))

table4ab <- pivot_wider(table1, id_cols = c("country", "continent"), names_from = "year", values_from = c("cases", "population"))

tmp <- mi_population
colnames(tmp)[3:4] <- c("population_2010", "population_2000")
mi_pop_tall <- pivot_longer(tmp, c("population_2010", "population_2000", "area_km2"), names_to = "measurement")

```


## Review: Loading data

- Format: **plain text, delimited**, proprietary, data bases
- Delimited files: lines are rows, columns separated by **delimiter** (like ",")
- Headers, column types
- Writing CSV files, RDS files

## Review: Handling dates

- Types: `Date` and `POSXIct` (date-time)
- Parsing: permutations of `ymd`, `%` keys
- Date-times, time zones
- Math and durations

## "Tidy" data, wide data

In this class we have focused on **tabular data**. Data organzied as **rows** and **columns**.

If we have a sample of $n$ items and have measured $k$ features of each item, our usual representation is:

- $k$ columns, one for each measurment
- $n$ rows, one for each unit

Our book calls this data "tidy." Our perspectives label this as "wide" data -- as we add more measurements, the table becomes wider.

## Tall formats

How else could we arrange data?

- One row per **value**
- Minimum two columns: "name" and "value"
- Additional columns to identify units

## Relational data

Sometimes we have multiple levels of data. For example,

- Clinics with patients
- Countries, measured over years
- Pitchers with many pitches thrown

It may be inefficient to duplicate upper level information for every lower level observation.

Solution: **relate** in two tidy tables

## Example: Tidy/Wide

```{r, echo = FALSE}

```
```{r}
table1
```


What are the units? What are the measurments?

## Example: Tall table
```{r echo = FALSE}
```
```{r}
table2
```


## Example: Relational (I)

```{r, echo= FALSE}
table3_countries <- group_by(table1, country) |> summarize(continent = first(continent))
table3_years <- tidyr::table1
```

```{r}
table3_countries 
table3_years
```


## Example: Relational (II)

```{r}
table4a # cases
table4b # population
```

## Why tall/untidy/relational data?

* Certain data sets just come that way
* Useful for plotting or summaries
* Space efficient for sparse data

## Exercise: summarizing on tall data

Recall our homework exercise to recreate R's built in `summary` function:

```{r}
summary(select(table1, cases, population))
```

Use `table2` to compute summaries per variable (just do mean, median, and max).

```{r}
table2
```

```{r summaryex, exercise = TRUE}

```

## Pivoting

If want to go from wide to tall or from tall to wide, we need to **pivot** our table.

* `pivot_longer`: takes a wide table and makes it tall
* `pivot_wider`: takes a tall table and makes it wide

## `pivot_longer`

Let's look at an abbreviated usage for [`pivot_longer`](https://tidyr.tidyverse.org/reference/pivot_longer.html):

```{r, eval = FALSE}
pivot_longer(
  data,
  cols,
  names_to = "name",
  values_to = "value",
  ...
)
```

>* `data`: a wide table
>* `cols`: the columns we want to place into the `value` column, the colnames will then populate the `names` column
>* `names_to` and `values_to` let you rename these columns if you wish


## Table 1 to Table 2: 

```{r}
colnames(table1)
pivot_longer(table1, c("cases", "population"))
```

**Note**: columns not included in list will be kept in original form (but duplicated per entry)

## More involved example from recent HW

```{r}
bbwater <- read_csv("../data/BKB_WaterQualityData_2020084.csv")
bbnumeric <- select(bbwater, where(is.numeric)) |> pivot_longer(everything())
bbnumeric
```

## Keeping other identifying columns

```{r}
bigwater <- pivot_longer(bbwater, where(is.numeric)) |>
  select(Site_Id, Read_Date, name, value)

bigwater
```

## Using in a plot

```{r}
ggplot(bigwater, aes(x = Site_Id, y = value)) + geom_boxplot() +
  facet_wrap(~ name, scales = "free")
```


## Untidy tall data

Sometimes we get data where **more than one observation** is placed on a row.

```{r}
table4a # cases
table4a # population
```
These are **country-year** observations.

## Using `pivot_longer` to fix

```{r}
cases <- pivot_longer(table4a, c("1999", "2000"), values_to = "cases", names_to = "year")
population <- pivot_longer(table4b, c("1999", "2000"), values_to = "population", names_to = "year")
cases
population
```

## Exercise

Pivot the following table to a tall format (what columns would make sense to go in `value`?). Would it make sense to use the `names_to` argument?

```{r}
mi_population 
```

```{r pivot_longer_ex, exercise = TRUE}

```

## Setting type of `name` column

In the previous we converted the years to strings. We probably want integers:

Method 1:
```{r}
pivot_longer(table4a, c("1999", "2000"), values_to = "cases", names_to = "year") |>
  mutate(year = as.integer(year))
```

Method 2:
```{r}
pivot_longer(table4a, c("1999", "2000"), values_to = "cases", names_to = "year",
             names_transform = as.integer)
```

## Prefixed columns

```{r, echo = FALSE}
table4a2 <- table4ab[1:4]
```

```{r}
table4a2
```

The `names_prefix` can remove the leading string:
```{r}
pivot_longer(table4a2, where(is.numeric), names_prefix = "cases_", 
             names_to = "year", values_to = "cases")
```




## More involved names

Another variation on tables we've seen so far looks like this:

```{r}
table4ab
```

```{r}
pivot_longer(table4ab, where(is.numeric), names_to = c("measurment", "year"), names_sep = "_")
```

## Going from tall tables to wide tables

When we have tables with too many rows, we can use [`pivot_wider`](https://tidyr.tidyverse.org/reference/pivot_wider.html) to restore one observation per row.

Abbreviated usage:
```{r, eval = FALSE}
pivot_wider(
  data,
  names_from = name,
  values_from = value,
  ...
)
```

>* `data` is a tall table
>* `names_from` is the `name` column of the tall table
>* `values_from` is the `value` column of the tall table

## `pivot_wider` example

```{r}
table2
pivot_wider(table2)
```

## Exercise 

Pivot the following table to an improved version.
```{r}
mi_pop_tall
```

```{r pwex, exercise = TRUE}

```

## Exercise continued

Continuing with `mi_pop_tall`, after moving population and area to own columns, use `pivot_longer` to get city-year measurements

```{r bigex1, exercise = TRUE}

```

