---
title: "Stats 306: Lecture 4"
subtitle: "Graphics: Facets, Coordinate Systems; Working with tables"
author: "Mark Fredrickson"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate) # install.packages("lubridate") if you don't have this
```

## Review: More on R

* Reviewed key concepts of graphics (visual variables) and `ggplot2` (creation, mapping)
* Showed documentation for available aesthetics, arguments, collections of functions
* Considered other geometries, multiple geometries
* Summary statistics: used by `geom_*` functions, can be used directly by us
* Output, R Graph Gallery

## Quantitative vs. Nominal measurements in R

We have will not overly emphasize the difference between **quantitative** (continuous, numerical) data and **nominal** (discrete, categorical) data *but* it's good to know how R handles these things.

```{r}
class(c(1.0, 2.2, 3.14159)) # quantitative
class(mpg$cyl) # quantitative
class(mpg$class) # nominal
class(factor(mpg$cyl)) # nominal
```

## `ggplot` and types of data

Some aesthetic mappings behave differently if they have quantitative or nominal columns.

By default, `cyl` will be treated as quantitative:
```{r}
ggplot(data = mpg, aes(x = hwy, y = cty, color = cyl)) +
  geom_jitter(size = 2)
```

## Forcing variables to be nominal

```{r}
ggplot(data = mpg, aes(x = hwy, y = cty, color = factor(cyl))) +
  geom_jitter(size = 2)
```

## Geometries that expect nominal mappings

```{r}
ggplot(data = mpg, aes(x = cyl, y = cty)) +
  geom_boxplot()
```

## What we probably wanted

```{r}
ggplot(data = mpg, aes(x = factor(cyl), y = cty)) +
  geom_boxplot()
```


## Overprinting revisited

We encountered **overprinting** when we were plotting cars' efficiency ratings. 

Our previous solution used jitter and/or transparency.

New idea: split up the plot by groups.

## Another overplotting example

```{r}
ggplot(data = mpg, aes(x = cty, y = hwy, color = class)) +
  geom_jitter()
```

## Introducing formulas

R has a special syntax for expressing **formulas**

```{r eval = FALSE}
y ~ x1 + x2 + x3 * x4
```

>* Used for specifying statistical models for Y given some predictors X
>* Delayed computation (early versions of `ggplot` used formulas in the `aes` function)
>* Anonymous functions in `tidyverse` packages

## Facets: "small multiples"

```{r}
ggplot(data = mpg, aes(x = cty, y = hwy, color = class)) +
  geom_jitter() +
  facet_wrap(~ class)
```

## Forcing on one row

```{r}
ggplot(data = mpg, aes(x = cty, y = hwy, color = class)) +
  geom_jitter() +
  facet_wrap(~ class, nrow = 1)
```

## Forcing two columns

```{r}
ggplot(data = mpg, aes(x = cty, y = hwy, color = class)) +
  geom_jitter() +
  facet_wrap(~ class, ncol = 2)
```

## Facets can include everything plots have 

```{r warning = FALSE}
ggplot(data = mpg, aes(x = cty, y = hwy, color = class)) +
  geom_jitter() +
  stat_smooth() + 
  facet_wrap(~ class) 
```


## Multiple factors

```{r warning = FALSE}
ggplot(data = mpg, aes(x = cty, y = hwy, color = class)) +
  geom_jitter() +
  stat_smooth() + 
  facet_wrap(~ class + drv) 
```

## Two way facets (`facet_grid`)

```{r warning = FALSE}
ggplot(data = mpg, aes(x = cty, y = hwy, color = class)) +
  geom_jitter() +
  stat_smooth() + 
  facet_grid(drv ~ class)
```

## Exercise

```{r facet-setup}
set.seed(293939222)
n <- 3 * 33
g <- sample(c("high", "medium", "low"), n, replace = TRUE) %>% as.factor()
xmu <- c(0, 1, 2)[g]
ymu <- c(0, -1, 2)[g]
x <- rnorm(n, mean = xmu, sd = sqrt(2)/2)
y <- rnorm(n, mean = ymu)
d <- data.frame(x, y, g)
```

Add facets using the `g` variable.
```{r facet, exercise = TRUE}
ggplot(data = d, aes(x = x, y = y, color = g)) +
  geom_point()
```

## Controlling the axis in `facet_grid`

We can make `facet_grid` behave like a single row (like a `facet_wrap`)

```{r warning = FALSE}
ggplot(data = mpg, aes(x = cty, y = hwy, color = class)) +
  geom_jitter() +
  facet_grid(. ~ class) 
```

## Going the other direction

Use the formula `class ~ .` to get a single column of facets:

```{r facet-column, exercise = TRUE}
ggplot(data = mpg, aes(x = cty, y = hwy, color = class)) +
  geom_jitter()
```




## Coordinate Systems

Thus far we have been using the default **Cartesian coordinate system**.

There are many other ways in which we can connect our **position visual variables** to their actual locations on a plane.

This are called **coordinate systems** or **projections**.

## Coordinate systems for presentation

Why use a different coordinate system? One reason to provide a convenient or interesting **presentation**.

Base plot:
```{r}
mpg_boxplot <- ggplot(data = mpg, aes(x = class, fill = class)) +
  geom_bar(width = 1) # width = 1 makes boxes touch
print(mpg_boxplot)
```

## Flipping coordinates

```{r}
mpg_boxplot + coord_flip()
```

## Coxcomb plots

Polar coordinates: express location as **angle** and **magnitude** from the origin.

```{r}
mpg_boxplot + coord_polar()
```

## Temperature data for Ann Arbor, MI

```{r}
# to run this line from the console, use `setwd("lectures")` first
aatemp <- read_csv("../data/ann_arbor_weather.csv.gz")
```

## Plotting temperature vs. time

```{r}
ggplot(data = aatemp, aes(x = DATE, y = TMAX)) +
  geom_line()
```

## Seasonal effects

We can see some variation in temperature over the year. What exactly is the **seasonal effect** of temperature? (Consistent variation over periods of time)
```{r}
# remember: tables are LISTS of VECTORS:
aatemp$month <- month(aatemp$DATE)
ggplot(data = aatemp, aes(x = factor(month), y = TMAX)) +
  geom_violin(aes(fill = factor(month)))
```

## Wrapping around using polar coordinates

```{r polar}
ggplot(data = aatemp, aes(x = factor(month), y = TMAX)) +
  geom_violin(aes(fill = factor(month))) +
  coord_polar()
```

## Other misceallaneous `ggplot` items

* Several other coordinate systems possible, in particular for represent geographic data.
* We saw using the `position` argument to jitter data. Other geometries have more options with jitter. See documentation.
* We'll revisit more with labels, themes, and other presentation later in the semester.

## Data formats: recap

We started working with data in the form of **tables**: (conceptually) lists of vectors.

```{r}
x <- 1:10
y <- 100:109
length(x)
length(y)
d <- list(column_x = x, column_y = y)
d
d$column_x
```

## Data formats: better types

A list of vectors could not guarantee they are all the same length. Instead we use **`data.frame`** and **`tibble`**.

```{r}
d2 <- as.data.frame(d)
dim(d2)
d3 <- as_tibble(d2)
d3
```

## Common operations on tables

* Pulling out a single column: `d$col`
* Creating a single column: `d$newcol <- EXPR`
* Pulling out several columns: `select(d, col1, col2)` (and several other forms)
* Creating new column (in new table): `mutate(d, newcol = f(col1, col2))`
* Pulling out rows: `filter(d, criteria)`
* Grouping and summarizing `group_by(d, discrete_column) |> summarize(new_name = a_function(col))`

## Getting and setting columns (one at a time)

Recall: to get something out of a list, use the dollar sign `$` operator

```{r}
d3$column_x
d3$x_plus_y <- d3$column_x + d3$column_y
head(d3, 2)
```

## Getting several columns at once

```{r}
select(d3, column_x, x_plus_y) |> head(2)
```

## Getting all but some columns

```{r}
select(d3, !column_x) |> head(2)
```

## Selectively grabbing columns

```{r}
d3$letter <- letters[d3$column_x]
d3$LETTER <- LETTERS[27 - d3$column_x]
head(d3, 2)
```
Get only "character" type columns:
```{r}
select(d3, where(is.character)) |> head(2)
```

Later, we'll introduce writing our own functions and see other things we can use instead of things like `is.character`.

## Grabbing rows by index

Tables in R allow for **two-dimensional indexing**: `tbl[rows, cols]`.

```{r}
d3[1, ]
d3[2:4, ]
d3[, c("column_x", "LETTER")]
d3[1:3, c("letter", "LETTER")]
```

## Grabbing rows by criteria (old school)

```{r}
d3[d3$column_y > 105, ]
```

## Grabbing rows (new school)

```{r}
filter(d3, column_y > 105)
```

## Multiple criteria

```{r}
filter(d3, column_y > 105 & column_x < 9)
```

```{r}
filter(d3, column_y > 108 | column_x  < 3)
```

## Exercise

Combine `filter` and `select` to get only the columns `cty` and `hwy` for cars that have more than 4 cylinders.

```{r}
head(mpg, 1)
```

```{r filter-select, exercise = TRUE}

```

## Next time

* Grouping
* Summarizing
* Creating new columns
* Writing our own functions
* R for DS: 5.6 - 5.7, 19.1 - 19.2
