---
title: "Stats 306: Lecture 8"
subtitle: "Wrapping up functions"
author: "Mark Fredrickson"
output: 
  slidy_presentation:
    incremental: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate) # install.packages("lubridate") if you don't have this
library(Lahman) # install.packages("Lahman") if you don't have this
```

## Review

* From duplicated to clean code:
  * Common patterns in function
  * Use of `mutate_at`
  * Lists of functions as arguments
* Return values: only one but can package up in vectors, lists, and tables
* Predicates: functions that return `TRUE` or `FALSE`; use in `select`
* Anonymous functions, tidyverse short hand notation `~ (.x - mean(.x))^2 / (length(.x) - 1)`

## The `Lahman` package

Sean Lahman maintains an extremely [detailed database of American baseball
statistics](https://www.seanlahman.com/baseball-archive/statistics). The
`Lahman` package conveniently imports this data for us.

```{r}
library(Lahman)
data(package = "Lahman")
dim(Batting)
colnames(Batting)
```

We can look up the description of the columns using `?Batting` or the help menu.

## Getting in the Lahmand data

```{r}
summarize(Batting, range(yearID))
group_by(Batting, playerID) |> 
  summarize_at(c("G", "HR", "RBI"), sum) |>
  arrange(desc(HR)) |> head(5)
```

## From IDs to full names

The `People` table has a `playerID` column. Does this uniquely identify players?

```{r}
dim(People)
summarize(People, n_distinct(playerID))
```

Looking up players: 
```{r}
lookupPlayer <- function(pid) {
  filter(People, playerID %in% pid) # result will be a table
}

lookupPlayer(c("aaronha01", "ruthba01")) |> select(nameFirst, nameLast)
```

## Successful players (batting)

Which players had the most home runs? Games played? Stolen bases?

```{r}
top_players <- function(stat) {
  group_by(Batting, playerID) |>
    summarize_at(stat, list(stat_sum = sum)) |>
    filter(stat_sum == max(stat_sum, na.rm = TRUE)) |>
    mutate(lookupPlayer(playerID)) |>
    select(playerID, stat_sum, nameFirst, nameLast) |>
    mutate(type = stat)
} 

# rbind stacks tables
rbind(
  top_players("HR"),
  top_players("G"),
  top_players("SB"))
```

(NB: using the `list(new_col_name = summary_function)` lets us force the name of the column with the results of `summary_function`)

## Variable number of arguments

Recall our template for creating a function in R:

```{r, eval = FALSE}
fname <- function(required_arg, optiona_arg = 123, ...) {
  # function body
}
```

The `...` means "any number of additional arguments"

## Example of variable arguments

```{r}
sum_summary <- function(...) {
  select(...) |> summarize_all(sum)
}

sum_summary(Batting, HR, G, SB)
```

Notice how all three arguments get passed to `select` automatically.


## Exercise

The `paste` function will take more than one item to turn into a string using its `...` argument:

```{r}
paste("Welcome", "to", "Stats206", "and", "DataSci101")
```

Write a function that takes a `...` argument and passes it along to `paste`. Set the `sep = "+"` when calling `paste`.

```{r exdots, exercise = TRUE}


## plus_paste("Welcome", "to", "Stats206", "and", "DataSci101")
```

## Common use of `...`: passing optional arguments to called functions

We have seen `na.rm = TRUE` pop up from time to time.

```{r}
d <- tibble(x = c(1, 10, NA, 7), y = c(1000, 2000, 10000, NA))
d
```

What happens if we try to summarize using `mean`?

```{r}
summarize_all(d, mean)
```
`summarize` uses `...` to pass additional arguments to the summarizing functions.

```{r}
summarize_all(d, mean, na.rm = TRUE)
```
## Fixing a previous function

```{r}
sum_summary <- function(...) {
  select(...) |> summarize_all(sum, na.rm = TRUE)
}

sum_summary(Batting, HR, G, SB)
```

## Conditional evaluation, short circuiting

We have used `if_else` to make choices for an entire vector. We often need **conditional program flow** in functions.

```{r}
my_median <- function(x) {
  x_sorted <- sort(x)
  n <- length(x)
  if (n %% 2 == 1) {
    ## odd numer of observations
    return(x_sorted[(n + 1)/2])
  } else {
    return(mean(x_sorted[c(n/2, n/2 + 1)]))
  }
  
  return(Inf) # this line will never run
}

values_even <- c(19, 2, 4, 11)
values_odd  <- c(19, 2, 4, 11, 9)

my_median(values_even)
my_median(values_odd)
```

## Careful about what you put in your `if` statements:

```{r eval = FALSE}

if (values_even > 3) {
  ... code might assume all values are larger than 3 ...
}
```
Would give the error message:
```
Error in if (values > 3) { : the condition has length > 1
```

Solution: use the `any` and `all` functions

## Exercise: Safe variance

Recall the usual (unbiased) estimator for variance:

$$\frac{\sum_{i = 1}^n (X_i - \bar X)^2}{n - 1}$$

If there is only one unit in the sample, this estimator is undefined. Instead, let's estimate the variance as 0 if we only have one observation.

Write a function to estimate variance in this way. Demonstrate on a vector of size 1 and a longer vector. Use `var` in the case where $n > 1$.

```{r}

```

## Combining more than one condition

We can use `&&` for "and" and `||` for "or". We can use `!` to negate.

Here's an example where we can only set one option for a function:

```{r}
food_order <- function(entree, side = NULL, dessert = NULL) {
  if (!is.null(side) && !is.null(dessert)) {
    return("You may only order one side or dessert")
  } else if (!is.null(side)) {
    return(paste("You ordered a main dish of", entree, "with a side of", side))
  } else if(!is.null(dessert)) {
    return(paste("You ordered a main dish of", entree, "with a dessert of", dessert))
  } else {
    # no side, no dessert
    return("Are you sure you wouldn't like to order a side or dessert?")
  }
}

food_order("soup")
food_order("soup", side = "salad")
food_order("soup", dessert = "cake")
food_order("soup", "salad", "cake")
```

## Next time

>* R for Data Science 7 - 7.8
