---
title: "Stats 306: Lecture 7"
subtitle: "Writing Functions"
author: "Mark Fredrickson"
output: 
  slidy_presentation:
    incremental: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate) # install.packages("lubridate") if you don't have this
aatemp <- read_csv("data/ann_arbor_weather.csv.gz")
```

## Review: `groupby` and `summarize`

* Use `groupby` to **partition** tables into groups
* Use `summarize` to compute **within group statistics**(e.g., conditional means)
* Useful functions to know about: `any`, `all`, `n`, `n_distinct`
* Interaction of `mutate` with `summarize`

## Review: creating functions

* Functions are variables like any other in R
* General syntax: `fname <- function(required, optional = 123) { # fn body }`
* Return will end a function and supply a value
* Last line of function is automatic return value

## Temperature data for Ann Arbor, MI

```{r}
aatemp
```

## From duplicated code to clean code

How many interquartile ranges above or below the median? (Similar to a Z-score)

```{r}
mutate(aatemp, TMAX_scale_IQR = (TMAX - median(TMAX)) / IQR(TMAX),
               TMIN_scale_IQR = (TMIN - median(TMIN)) / IQR(TMIN)) |>
  group_by(abs(TMAX_scale_IQR) > 1.5, abs(TMIN_scale_IQR) > 1.5) |>
  summarize(n())
```

## Writing our own functions: capturing common behavior

```{r}
scale_IQR <- function(x) {
  (x - median(x)) / IQR(x)
}

temp_summary <- function(tbl) {
  group_by(tbl, abs(TMAX_scale_IQR) > 1.5, abs(TMIN_scale_IQR) > 1.5) |>
  summarize(n())
}

mutate(aatemp, TMAX_scale_IQR = scale_IQR(TMAX), 
               TMIN_scale_IQR = scale_IQR(TMIN)) |>
  temp_summary()
```

## Even better: `mutate_at`

We're still calling the function twice. `mutate_at` lets us pass in a
**function** as an argument!

If we give one function, it uses the original column names:

```{r}
mutate_at(aatemp, c("TMAX", "TMIN"), scale_IQR) |>
  group_by(abs(TMAX) > 1.5, abs(TMIN) > 1.5) |>
  summarize(n())
```

## More than one function, giving them names

We can give a list of more than one function of the form
```{r eval = FALSE}
list(name1 = function1, name2 = function2)
```

```{r}
mutate_at(aatemp, c("TMAX", "TMIN"), 
          list(scale_IQR = scale_IQR, scale = scale)) |>
  colnames()
            
```

## Exercise

**Skew** measures the relative amount of variation above and below the center of distribution. The sample **coefficient of skew** is given by 

$$ \frac{(1/n) \sum_{i=1}^n (X_i - \bar X_n)^3}{\hat \sigma^3}$$
where $\hat \sigma$ is the sample standard deviation.

Write a function to compute the coefficient of skew and use `summarize_at` to compute it for `TMAX` and `TMIN` in the `aatemp` data set.

```{r coefskew, exercise = TRUE}

```

## Plot of `TMAX` and `TMIN`

```{r}
tall_temp <- select(aatemp, TMAX, TMIN) |> pivot_longer(c("TMAX", "TMIN"))
tall_temp
ggplot(tall_temp, aes(x = value, fill = name)) + geom_density(alpha = 0.5)
```

## Return values

R functions can return (at most) one object. Can we return several items? Yes: using vectors, lists, and tables.

```{r}
first_last <- function(x) { c(x[1], x[length(x)]) }
first_last(LETTERS)

mean_range <- function(x) { list(mean = mean(x), range = range(x))}
mr <- mean_range(rnorm(100)) # 100 random values
mr$mean
mr$range

numbered_table <- function(x) { 
  tibble(idx = seq_along(x), main_column = x)
}
numbered_table(letters)
```


## Summarizing with functions that return more than one value

```{r}
mean_var <- function(x) {
  tibble(average = mean(x), variance = var(x))
}
aat_year <- group_by(aatemp, year(DATE))
summarize(aat_year, mean_var(TMAX))
```

Use caution when using this function more than once!

```{r}
summarize(aat_year, mean_var(TMAX), mean_var(TMIN))
```

## Predicates

A **predicate** is a function that returns `TRUE` or `FALSE`.

For example:
```{r}
no_missing <- function(x) {
  !(any(is.na(x)))
}

no_missing(c(2,3,44))
no_missing(c(7, NA, 122))
```

## Selecting columns with predicates

We have seen several predicates used already: `is.numeric` or `is.character` in selecting columns. We can also use our own predicates.

```{r}
select(aatemp, where(no_missing)) |> colnames()
```

## Anonymous functions

Sometimes we want to write short little functions that only get used once. We call these **anonymous functions** like this:

```{r}
select(aatemp, where(function(x) { !any(is.na(x)) })) |> ncol()
```

## Tidyverse anonymous function short hand

Many of the tidyverse functions support using formulas to indicate an anonymous function:

```{r}
select(aatemp, where(~ !any(is.na(.x)))) |> ncol()
```

>* `~` starts a formula
>* `.x` and `.y` are the arguments to the function

We'll see more of this form when we discuss iteration.

## More function short hand

```{r}
summarize_at(aatemp, c("TMAX", "TMIN"), ~ mean((.x - mean(.x))^3) / sd(.x)^3)
```

## Functions with optional arguments

Recall that R functions with `argument = value` are optional. We can write functions this way too.

```{r}

multSomething <- function(x, to_mult = 1) {
  x * to_mult
}

multSomething(10)
multSomething(10, 2)
multSomething(to_mult = 2, 10)

mutate(aatemp, TMAX1 = multSomething(TMAX),
               TMAX2 = multSomething(TMAX, 2)) |> 
  select(TMAX, TMAX1, TMAX2)
```

## Some functions allow passing additional arguments indirectly

```{r}
mutate_at(aatemp, c("TMAX", "TMIN"), multSomething, to_mult = 0) |>
  select(TMAX, TMIN)
```

This is done using the `...` argument to `mutate_at`.

## Exercise

Write a function `center` that **takes a function** as an argument computes
$$x - f(x)$$
Make the default argument be `mean`, but demonstrate using `median` centering as well.

```{r higher-order, exercise = TRUE}
myvec <- c(203404, 292, 1010, 3, -10930, 39)
```

## Exercise

The `paste` function will take more than one item to turn into a string using its `...` argument:

```{r}
paste("Welcome", "to", "Stats206", "and", "DataSci101")
```

Write a function that takes a `...` argument and passes it along to `paste`. Set the `sep = "+"` when calling `paste`.

```{r exdots, exercise = TRUE}


## plus_paste("Welcome", "to", "Stats206", "and", "DataSci101")
```

## Conditional evaluation, short circuiting

We have used `if_else` to make choices for an entire vector. We often need **conditional program flow** in functions.

```{r}
my_median <- function(x) {
  x_sorted <- sort(x)
  n <- length(x)
  if (n %% 2 == 1) {
    ## odd numer of observations
    return(x_sorted[(n + 1)/2])
  } else {
    return(mean(x_sorted[c(n/2, n/2 + 1)]))
  }
  
  return(Inf) # this line will never run
}

values_even <- c(19, 2, 4, 11)
values_odd  <- c(19, 2, 4, 11, 9)

my_median(values_even)
my_median(values_odd)
```

## Careful about what you put in your `if` statements:

```{r eval = FALSE}

if (values_even > 3) {
  ... code might assume all values are larger than 3 ...
}
```
Would give the error message:
```
Error in if (values > 3) { : the condition has length > 1
```

Solution: use the `any` and `all` functions

## Exercise: Safe variance

Recall the usual (unbiased) estimator for variance:

$$\frac{\sum_{i = 1}^n (X_i - \bar X)^2}{n - 1}$$

If there is only one unit in the sample, this estimator is undefined. Instead, let's estimate the variance as 0 if we only have one observation.

Write a function to estimate variance in this way. Demonstrate on a vector of size 1 and a longer vector. Use `var` in the case where $n > 1$.

```{r}

```

## Combining more than one condition

We can use `&&` for "and" and `||` for "or" (exclusive). We can use `!` to negate.

Here's an example where we can only set one option for a function:

```{r}
food_order <- function(entree, side = NULL, dessert = NULL) {
  if (!is.null(side) && !is.null(dessert)) {
    return("You may only order one side or dessert")
  } else if (!is.null(side)) {
    return(paste("You ordered a main dish of", entree, "with a side of", side))
  } else if(!is.null(dessert)) {
    return(paste("You ordered a main dish of", entree, "with a dessert of", dessert))
  } else {
    # no side, no dessert
    return("Are you sure you wouldn't like to order a side or dessert?")
  }
}

food_order("soup")
food_order("soup", side = "salad")
food_order("soup", dessert = "cake")
food_order("soup", "salad", "cake")
```

## Next time

>* R for Data Science 7 - 7.8
