---
subtitle: "Stats 306: Lecture 22 (Rcpp)"
title: "Rcpp"
author: "Mark Fredrickson"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate)
library(Rcpp)
```

## Rcpp: easy R interface

R has always had a C/C++ interface, but it is tricky to use. 

* You need to package up your data into specific forms
* You need to unpack results back into R native formats
* Does not interact well with calling R functions from C/C++
* You need to manage all compilation and packaging of code

`Rcpp` is a library of convenience functions and C++ templates that greatly eases these burdens. It also has facilities to make writing C++ seem more like writing R.

(Note: you need to install C/C++ tool chain. [See 25.1 in Advanced R](https://adv-r.hadley.nz/rcpp.html#prerequisites-17))

## Example Rcpp usage

The `cppFunction` takes a string that is valid C and turns it into a function:

```{r}
cppFunction('int add(int x, int y, int z) {
  int sum = x + y + z;
  return sum; // notice no parens
}')

# now `add` defined:
add(1, 2, 3)
```

Some differences with C/C++

* All lines must end with `;`
* Functions must explicitly `return` (but you don't need parens)
* Variables must be declared and typed
* All variables are **scalar** by default, but Rcpp includes some convenient **collections**
* Comments given by `//` or `/* ... multiple lines...  */` blocks

## Collections and Loops

```{r}
cppFunction('double sumC(NumericVector x) {
  int n = x.size(); // 
  double total = 0;
  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total;
}')

sumC(1:10) 
```

* Notice `x` is marked as `NumericVector` object
* `x.size()` is **method** call. Methods are like functions `f(obj, arg2)` but `obj.method(arg2)`.
* Loops have a different syntax `(create var; until condition; increment var)`
* C/C++ counts start at 0 instead of 1 (e.g. `x[0]` is first item in collection)

## Comparing performance

```{r}
sumR <- function(x) { reduce(x, `+`) }
bench::mark(sumR(1:100), sumC(1:100), sum(1:100))
```
## Better iteration

```{r}
cppFunction('double sumC2(NumericVector x) {
  double total = 0;
  NumericVector::iterator it;
  for(it = x.begin(); it != x.end(); ++it) {
    total += *it; // "dereference" the point `it`
  }
  return total;
}')

sumC2(1:100)
```

## Outer products

Recall that for vectors $x$ and $y$, the outer product $M$ is a matrix with $M_{ij} = x_i y_j$.

```{r}
outerR <- function(x, y) {
  m <- matrix(0, nrow = length(x), ncol = length(y))
  for (j in 1:length(y)) {
    m[, j] <- y[j] * x
  }
  return(m)
}
a <- 1:10
b <- c(20, 30, 50)
outerR(a, b)
v1 <- rnorm(1000)
v2 <- rnorm(1000)
bench::mark(outer(v1, v2), outerR(v1, v2))
```

## Implementing in Rcpp

Go to RStudio...

## Using `.cpp` files

When writing code in `.cpp` files we can either compile it directly or have it included in a package definition (see `vignette("Rcpp-package")`) or use `sourceCpp`:

```{r}
sourceCpp("lecture22_outerC.cpp")
outerC(a,b)

bench::mark(outerR(v1, v2), outer(v1, v2), outerC(v1, v2))
```

## Rcpp other notes

* Usually write in R first, then translate to C++
* Great candidate for functions that call themselves (recursive functions) or otherwise have deep call stacks
* Opportunities to use linear algebra systems in C++ (RcppAmarmadillo)
* Wide variety of interesting data types in the Standard Template Library, Boost, other packages
* Parallelization is more difficult; better to keep parallel techniques at the R level

