---
title: "Stats 306: Lab 5"
subtitle: "Midterm Review"
author: "Your Name"
output: 
  slidy_presentation:
    incremental: true
runtime: shiny_prerendered
---

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# This just checks if students need to install some packages that they might
# not have.
if (!require(gapminder)) install.packages("gapminder", repos = "http://cran.us.r-project.org")

if (!require(nycflights13)) install.packages("nycflights13")
if (!require(imager)) install.packages("imager")
```

```{r setup, echo=FALSE, include=FALSE}
library(learnr) # For interactive exercises
library(tidyverse) # For ggplot2, dplyr
library(gapminder)
library(nycflights13)
library(imager)
```

## Logistics

> - Midterm one(1) to be released on 10/05 and due on 10/07.
> - Materials covered up to the week of September 26 (last week). 



## Today's Content

> - Today's lab combines both conceptual problems and coding
> - Covers topics covered until last week: (ggplot, data manipulation...)


## Problem 1

> - Take a glimpse at the first six rows of the `mpg` data.
> - What happens if you make a scatterplot of `class` vs `drv`? Why is the plot not useful?

```{r mpg, exercise = T}
head(mpg)
```

```{r classVsdrv,  exercise = T}
ggplot(mpg, aes(x = class, y = drv)) +
  geom_point()
```

## Answer: 
The resulting scatterplot has only a few points.
A scatter plot is not a useful display of these variables since both `drv` and `class` are categorical variables. Since categorical variables typically take a small number of values, there are a limited number of unique combinations of `(x, y)` values that can be displayed. In this data, `drv` takes 3 values and `class` takes 7 values, meaning that there are only 21 values that could be plotted on a scatterplot of `drv` vs. `class`. In this data, there 12 values of (drv, class) are observed. Scatter plots are prefered for visualizing two continuous variables.



## Problem 2

> - What’s gone wrong with this code? Why are the points not blue?
> - How would you fix this problem (use the code cell below)? 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```


```{r aesthetics, exercise = T}
## fix the problem
```

## Answer:
The argument `colour = "blue"` is included within the mapping argument, and as such, it is treated as an aesthetic, which is a mapping between a variable and a value. To fix this, keep color out of the aesthetics attribute.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```


## Problem 3
 > - What happens if you facet on a continuous variable?
 > - Use the plot below as a hint.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(. ~ cty)
```


## Answer:

The continuous variable is converted to a categorical variable, and the plot contains a facet for each distinct value.

## Problem 4
> - Discuss how to obtain each of the plot below

```{r fig}
myimg <- load.image("fig.png")
plot(myimg, axes = FALSE)
```


> - Recreate the R code necessary to generate the following graphs.

```{r fig1, exercise = T}
## Figure 1
```

```{r fig2, exercise = T}
## Figure 2
```

```{r fig3, exercise = T}
## Figure 3
```


```{r fig4, exercise = T}
## Figure 4
```


```{r fig5, exercise = T}
## Figure 5
```


```{r fig6, exercise = T}
## Figure 6
```



## Answer

```{r fig1_s, exercise = T}
## Figure 1
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(se = FALSE)
```



```{r fig2_s, exercise = T}
## Figure 2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(group = drv), se = FALSE) +
  geom_point()
```


```{r fig3_s, exercise = T}
## Figure 3
ggplot(mpg, aes(x = displ, y = hwy, colour = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```


```{r fig4_s, exercise = T}
## Figure 4
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(colour = drv)) +
  geom_smooth(se = FALSE)
```

```{r fig5_s, exercise = T}
## Figure 5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(colour = drv)) +
  geom_smooth(aes(linetype = drv), se = FALSE)
```


```{r fig6_s, exercise = T}
## Figure 6
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(size = 4, color = "white") +
  geom_point(aes(colour = drv))
```

## Problem 5
> - We have learned that the `filter()` is used to  subset observations (rows) based on their values.
> - Consider the flight data:

```{r flights}
head(flights, n = 3)
str(flights)
```

> - How would you find all flights that:
> - Had an arrival delay of two or more hours (need to convert time from hrs to mins)? ***[Hint: look up the `arr_delay` variable.]***

```{r arrival, exercise = T}
## your answer here
```

> -  Flew to Houston (IAH or HOU)? ***[Hint: look up the `dest` variable]***

```{r destination, exercise = T}
## your answer here
```

> -  Were operated by United (UA), American (AA), or Delta (DL)? ***[Hint: look up the `carrier`]***

```{r airlines, exercise = T}
## your answer here
```

## Answer:
> -  An arrival delay of two or more hours

```{r arrival_sol, exercise = T}
filter(flights, arr_delay >= 120)
```


> -  Flew to Houston (IAH or HOU)?

```{r destination_sol, exercise = T}
filter(flights, dest == "IAH" | dest == "HOU")


## Or Preferably
filter(flights, dest %in% c("IAH", "HOU"))
```

> - Were operated by United, American, or Delta

```{r airlines_sol, exercise = T}
filter(flights, carrier %in% c("AA", "DL", "UA"))
```


## Problem 6
> - We are interested in selecting the following columns:
> - `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from the `flights` data.
> - John proposes the use of 
```
select(flights, all_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
```
> - while Jane proposes:
```
select(flights, any_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
```

> - They are unsure if they will get same answers. Without running any of the codes, do you think they will get same answers? Explain.

## Answer:
> - Yes, they will get the same answers. The difference between the two functions is that, `all_of()` throws and error when a variable is missing in the dataframe while `any_of()` does not. 

> - The example codes confirms that they yield same output.

```{r all_of, exercise = T}
select(flights, all_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
```

```{r any_of, exercise = T}
select(flights, any_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
```


> - Now compare the two codes below
> - Here, I add a variable that is not in the `flights` data -- "`petal`"
> - Note that `all_of()` breaks completely
while `any_of()` selects those columns that exist in the dataset.

```{r all_of_err, exercise = T}
select(flights, all_of(c("dep_time", "dep_delay", "arr_time", "arr_delay", "petal")))
```

```{r any_of_err, exercise = T}
select(flights, any_of(c("dep_time", "dep_delay", "arr_time", "arr_delay", "petal")))
```


## Problem 7
> - Sam wants to compute the sample skewness defined by.

$$
Skew(x) = \dfrac{\frac{1}{n-1}( \sum_{i = 1}^{n}(x_i - \bar{x})^3)}{Var(x)^{3/2}}
$$

> - She writes her codes as follows, following the DRY (Do Not repeat Yourself) principle

```
skewness <- function(x) {
  n  <- length(x)
  m  <- mean(x)
  v  <- var(x)
  dev_cub <- (x - m) ^ 3
  (sum(dev_cub) / (n - 2)) / v ^ (3 / 2)
}
```
> - She is confident that her functions works and tries it on

```
x <- c(4, 5, 8, 1, 4 , 7 , 8)
```

> - Jesse thinks something might go wrong if he supplied the following data

```
y <- c(5, 6, 2, 9, NA)
```

> - Is Jesse's claim true? Why would it work or not without running Sam's codes?


> - if you agrre with Jesse, how would you fix this problem?

```{r Sam, exercise = T}
#### Your code here
```



## Answer:
Jesse is right! The function breaks on his data because there is an `NA`. Since `NA` is not a number, we cannot take the average of the vector (you cannot sum numbers and NAs). A solution is to allow for the `na.rm` argument in the function.

```{r Sam_sol, exercise = T}
#### Fix
skewness <- function(x, na.rm = FALSE) {
  n <- length(x)
  m <- mean(x, na.rm = na.rm)
  v <- var(x, na.rm = na.rm)
  dev_cub <- (x - m) ^ 3
  (sum(dev_cub, na.rm = na.rm) / (n - 2)) / v ^ (3 / 2)
}

y <- c(5, 6, 2, 9, 7, 2, 5, NA)
skewness(y, na.rm = TRUE)
```

