---
title: "Stats 306: Lab 3"
subtitle: "Graphics: Facets, Coordinate Systems; Working with tables"
author: "Your Name"
output: 
  slidy_presentation:
    incremental: true
runtime: shiny_prerendered
---

## Set up

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# This just checks if students need to install some packages that they might 
# not have.

if (!require(ggforce)) install.packages("ggforce")
if (!require(maps)) install.packages("maps")
if (!require(mapproj)) install.packages("mapproj")
if (!require(lattice)) install.packages("lattice")
```

```{r setup, echo=TRUE}
library(learnr)     # For interactive exercises
library(tidyverse)  # For ggplot2, dplyr
library(maps)       # For plotting maps 
library(ggforce)    # For hurricane map, later. 
```

## Logistics

> - HW / Quiz dates 
> - Questions?  

## Today's Content

* `ggplot2` facets.
* `ggplot2` coordinate systems.
* Table operations and `dplyr` verbs. 

## Facets 

Faceting allows us to split data into subset and plot similar graphs for each subset. 
In lecture we saw that facets are often very useful to avoid over-fitting. 
There are three types of facets in `ggplot2`: 

* `facet_wrap`: Creates a one-dimensional set of plots, that are often shown in a 2d grid.
* `facet_grid`: Creates a two-dimensional grid of panels, defined by two different columns. 
* `facet_null`: This is the default facet, which is just a single plot

## Difference between `facet_wrap` and `facet_grid`

\
\
\

<center>

![A sketch difference between facetting systems. Figure borrowed from [ggplot2 book](https://ggplot2-book.org/facet.html)](images/facets.png)

</center>

## `facet_wrap` Example

## `facet_grid` Example

## `facet_wrap` Exercise

For this exercise, we are going to use the `barley` data from the `lattice` package. 
This is famous statistical dataset that has been cited in many papers.
We will use it to create a *Cleveland dot plot*. 

```{r}
barley <- lattice::barley
head(barley)
```

**Exercise**: Add a facets and colors to the following code to reduce over-plotting. 
Do you notice anything strange about the data? 

```{r}
ggplot(barley) + 
  geom_point(aes(x = yield, y = variety, col = year)) + 
  facet_wrap(~site) + 
  theme_bw()
```

## `facet_wrap` Solution 

William Cleveland (to whom the term "data science" is often attributed) famously made a plot similar to the following in a textbook he published:

```{r}
ggplot(barley) + 
  geom_point(aes(x = yield, y = variety, col = year)) + 
  facet_wrap(~site) + 
  theme_bw()
```

In this plot, he points out that an oddity in Morris: it's the only site in which the 1932 crop yield was higher than that in 1931! 
Using this plot, he claimed that the data must have been mislabeled (although there is [some recent argument](https://blog.revolutionanalytics.com/2014/07/theres-no-mistake-in-the-barley-data.html) that this is not the case). 

## Coordinate Systems

In our last lecture, we looked closely at various *coordinate systems* in `ggplot2`. 
Coordinate systems allow us to easily change how the data are displayed, with limited effort. 

There are *linear* coordinate systems that don't change the shape of `geoms`: 

> - `coord_cartesian()`: Default coordinate system
> - `coord_flip()`: Default coordinate system, but flipping `x` and `y` axis. 
> - `coord_fixed()`: Default coordinate system, but fixing the ratio between `x` and `y` axis. 

There are also *non-linear* coordinate systems, that can change the shape of `geoms`: 

> - `coord_polar()`: Polar coordinates.
> - `coord_map()`: Map projections (spherical earth to 2D plane).
> - `coord_trans()`: Custom transformations to `x` and `y` positions. 

## Polar Coordinate Example

## Polar Coordinate Exercise

## `coord_map()` Example

Let's take a look at a new coordinate system, using `coord_map()`. 
First, we will use the `maps` package to get information about some North American Countries. 

```{r NorthAmerica, echo=TRUE}
world <- map_data("world") %>%
  filter(region %in% c("USA", "Canada", "Mexico", "Jamaica",
                       "Cuba", "Panama", "Guatemala", "Puerto Rico", 
                       "Costa Rica", "Dominican Republic", "Haiti",
                       "El Salvador", "Belize", "Bahamas", "Nicaragua",
                       "Honduras", "Trinidad")) %>%
  filter(long < 0)
```

We can plot this data using `ggplot2` and `geom_polygon()`. 
Lets save the map to use later:

```{r plotNorthAmerica, echo=TRUE}
my_map <- ggplot(world) + 
  geom_polygon(aes(x = long, y = lat, group = group), fill = 'white',
               col = 'black') + 
  theme_bw()

print(my_map)
```

## `coord_map()` continued

The problem with the previous slide is that the polygons are a bit distorted, so the polygons don't look exactly like we would expect.
`coord_map()` can help fix this issue.
In particular, we will start with `coord_quickmap()`, which is the simplest implementation. 

```{r coordQuickMap, echo=TRUE}
my_map + 
  coord_quickmap()
```

## `coord_map()` Continued

There are probably better ways to project the 3D world onto a 2D surface. 
You can explore some of these options by looking at the help page for `coord_map` and `mapproj::mapproject`.
Let's use the [globular](https://en.wikipedia.org/wiki/Nicolosi_globular_projection) projection

```{r}
my_map + 
  coord_map('globular')
```

## Hurricane Map

```{r}
my_storms <- storms %>%
  filter(year == 2016, name %in% c('Matthew', "Nicole", "Karl", "Otto"))

my_map + 
  coord_map('globular') + 
  geom_path(data = my_storms, aes(x = long, y = lat), col = 'red') +
  geom_circle(
    data = my_storms, 
    aes(x0 = long, y0 = lat, r = tropicalstorm_force_diameter/120),
    col = 'red'
  ) +
  coord_map('globular') +  # results in distorted storm circles
  guides(color = "none") + 
  theme_bw() + 
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  ) + 
  facet_wrap(~name) + 
  ggtitle("Selected 2016 Tropical Storms and Hurricanes")
```

## Table operations

Today we are going to focus on functions in the `dplyr` package, which is part of the `tidyverse`. 
The `dplyr` package aims to provide a function for each basic "verb" of data manipulation. 
Here are some common functions / "verbs" for data manipulation: 

* Rows: 
    + `filter()`: Select rows based on criteria on column values
    + `slice()`: choose rows based on locations.
    + `arrange()`: change the order of the rows.
* Columns: 
    + `select()`: Pick and choose columns to include in data.
    + `pull()`: Select a single column (similar to `data$column`).
    + `rename()`: Change the name of columns.
    + `mutate()`: Change the values of columns / creates new columns. 
* Groups (of rows):
    + `group_by()`: Creates a group of rows.
    + `summarise()` or `summarize()`: Calculate summary values for a given group. 

## `mutate()`, `filter()`, `select()` Examples

## `mutate()`, `filter()`, `select()`, `arrange()` Exercise

## `group_by()` and `summarize()` (1-2 slides)

Introduce the function `n = n()`

## `group_by()` and `summarize()` examples. 
