---
title: "Stats 306: Lab 3"
subtitle: "Graphics: Facets, Coordinate Systems; Working with tables"
author: Yilei Zhang
output: 
  slidy_presentation:
    incremental: true
runtime: shiny_prerendered
---

## Set up

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# This just checks if students need to install some packages that they might 
# not have.

if (!require(ggforce)) install.packages("ggforce", repos = "http://cran.us.r-project.org")
if (!require(maps)) install.packages("maps", repos = "http://cran.us.r-project.org")
if (!require(mapproj)) install.packages("mapproj", repos = "http://cran.us.r-project.org")
if (!require(lattice)) install.packages("lattice", repos = "http://cran.us.r-project.org")
```

```{r setup, echo=TRUE}
library(learnr)     # For interactive exercises
library(tidyverse)  # For ggplot2, dplyr
library(maps)       # For plotting maps 
library(ggforce)    # For hurricane map, later. 
```

## Logistics

> - HW 1 is due on September 25
> - HW 2 is due on October 2
> - Quiz 2 is due on September 25

> - Questions?  

## Today's Content

* `ggplot2` facets.
* `ggplot2` coordinate systems.
* Table operations and `dplyr` verbs. 

## Facets 

Faceting allows us to split data into subset and plot similar graphs for each subset. 
In lecture we saw that facets are often very useful to avoid over-fitting. 
There are three types of facets in `ggplot2`: 

* `facet_wrap`: Creates a one-dimensional set of plots, that are often shown in a 2d grid.
* `facet_grid`: Creates a two-dimensional grid of panels, defined by two different columns. 
* `facet_null`: This is the default facet, which is just a single plot

## Difference between `facet_wrap` and `facet_grid`

\
\
\

<center>

![A sketch difference between facetting systems. Figure borrowed from [ggplot2 book](https://ggplot2-book.org/facet.html)](images/facets.png)

</center>

## `facet_wrap` Example

Lets use the `ToothGrowth` dataset:

```{r headTooth, exercise=TRUE}
head(ToothGrowth)
```

```{r uniqueDose, exercise=TRUE}
unique(ToothGrowth$dose)
```

A box-plot of all of the data: 

```{r}
tooth_gg <- ggplot(ToothGrowth) + 
  geom_boxplot(aes(y = len)) + 
  theme_bw()

print(tooth_gg)
```


```{r createBasicBox, echo=TRUE, include=FALSE}
tooth_gg <- ggplot(ToothGrowth) + 
  geom_boxplot(aes(y = len)) + 
  theme_bw()
```

Now if we want to summarize at the distribution for tooth growth for each supplement, we can facet on supplement: 

```{r wrappedBox, exercise=TRUE, exercise.setup="createBasicBox"}
tooth_gg + 
  facet_wrap(~supp)
```

If we wanted to have two rows of figures instead of two columns, we can specify this with `ncol = 1` or `nrow = 2`

```{r wrappedBox2, exercise=TRUE, exercise.setup="createBasicBox"}
tooth_gg + 
  facet_wrap(~supp, ncol = 1)
```

We can get the same figure (but with labels on the side) by using `facet_grid`.

```{r gridBox1, exercise=TRUE, exercise.setup="createBasicBox"}
tooth_gg + 
  facet_grid(supp~.)
```

In the previous code, we are saying that we want the rows of the grid to correspond to the variable `supp`, and we want to give it an empty variable `.` for the columns. 
We can "rotate" the figure by switching the order of these variables: 

```{r gridBox2, exercise=TRUE, exercise.setup="createBasicBox"}
tooth_gg + 
  facet_grid(.~supp)
```

We can of course include a second variable rather than a missing variable if we wanted to do so: 

```{r gridBox3, exercise=TRUE, exercise.setup="createBasicBox"}
tooth_gg + 
  facet_grid(supp~dose)
```

The order of the variables depends on what information we are most interested in showing.
For example, in the previous plot, the order makes it easiest to compare the effect of various different levels of dose for each supplement. 
If we were interested how supplements compare for similar dosages, the following plot may be preferred:

```{r gridBox4, exercise=TRUE, exercise.setup="createBasicBox"}
tooth_gg + 
  facet_grid(dose~supp)
```

## `facet_wrap` Exercise

For this exercise, we are going to use the `barley` data from the `lattice` package. 
This is famous statistical dataset that has been cited in many papers.
We will use it to create a *Cleveland dot plot*. 

```{r}
barley <- lattice::barley
head(barley)
```

```{r loadBarley}
barley <- lattice::barley
```

**Exercise**: Add a facets and colors to the following code to reduce over-plotting. 
Do you notice anything strange about the data? 

```{r facetExercise, exercise=TRUE, exercise.setup="loadBarley"}
ggplot(barley) + 
  geom_point(aes(x = yield, y = variety)) + 
  theme_bw()
```

## `facet_wrap` Solution 

William Cleveland (to whom the term "data science" is often attributed) famously made a plot similar to the following in a textbook he published:

```{r}
ggplot(barley) + 
  geom_point(aes(x = yield, y = variety, col = year)) + 
  facet_wrap(~site) + 
  theme_bw()
```

In this plot, he points out that an oddity in Morris: it's the only site in which the 1932 crop yield was higher than that in 1931! 
Using this plot, he claimed that the data must have been mislabeled (although there is [some recent argument](https://blog.revolutionanalytics.com/2014/07/theres-no-mistake-in-the-barley-data.html) that this is not the case). 

## Coordinate Systems

In our last lecture, we looked closely at various *coordinate systems* in `ggplot2`. 
Coordinate systems allow us to easily change how the data are displayed, with limited effort. 

There are *linear* coordinate systems that don't change the shape of `geoms`: 

> - `coord_cartesian()`: Default coordinate system
> - `coord_flip()`: Default coordinate system, but flipping `x` and `y` axis. 
> - `coord_fixed()`: Default coordinate system, but fixing the ratio between `x` and `y` axis. 

There are also *non-linear* coordinate systems, that can change the shape of `geoms`: 

> - `coord_polar()`: Polar coordinates.
> - `coord_map()`: Map projections (spherical earth to 2D plane).
> - `coord_trans()`: Custom transformations to `x` and `y` positions. 

## Polar Coordinate Example

One of the most common coordinate transformations we see for data visualizations is **Polar-Coordinates**. 
The idea for polar coordinate mapping is that sometimes displaying the information as an angle and radius is more informative than displaying the information in the `x, y` Cartesian plane,
for example polar coordinates often help us visualize seasonal effects. 

```{r mtcarsPolar, exercise=TRUE}
my_cars <- mtcars
my_cars$model <- rownames(mtcars)
ggplot(my_cars) + 
  geom_bar(aes(y = mpg, x = model), stat = 'identity') + 
  coord_polar() + 
  theme_bw()
```

## `coord_map()` Example

Let's take a look at a new coordinate system, using `coord_map()`. 
First, we will use the `maps` package to get information about some North American Countries. 

```{r NorthAmerica, echo=TRUE}
world <- map_data("world") %>%
  filter(region %in% c("USA", "Canada", "Mexico", "Jamaica",
                       "Cuba", "Panama", "Guatemala", "Puerto Rico", 
                       "Costa Rica", "Dominican Republic", "Haiti",
                       "El Salvador", "Belize", "Bahamas", "Nicaragua",
                       "Honduras", "Trinidad")) %>%
  filter(long < 0)
```

We can plot this data using `ggplot2` and `geom_polygon()`. 
Lets save the map to use later:

```{r plotNorthAmerica, echo=TRUE}
my_map <- ggplot(world) + 
  geom_polygon(aes(x = long, y = lat, group = group), fill = 'white',
               col = 'black') + 
  theme_bw()

print(my_map)
```

## `coord_map()` continued

```{r preLoadMap}
world <- map_data("world") %>%
  filter(region %in% c("USA", "Canada", "Mexico", "Jamaica",
                       "Cuba", "Panama", "Guatemala", "Puerto Rico", 
                       "Costa Rica", "Dominican Republic", "Haiti",
                       "El Salvador", "Belize", "Bahamas", "Nicaragua",
                       "Honduras", "Trinidad")) %>%
  filter(long < 0)

my_map <- ggplot(world) + 
  geom_polygon(aes(x = long, y = lat, group = group), fill = 'white',
               col = 'black') + 
  theme_bw()
```

The problem with the previous slide is that the polygons are a bit distorted, so the polygons don't look exactly like we would expect.
`coord_map()` can help fix this issue.
In particular, we will start with `coord_quickmap()`, which is the simplest implementation. 

```{r coordQuickMap, exercise=TRUE, exercise.setup="preLoadMap"}
my_map + 
  coord_quickmap()
```

## `coord_map()` Continued

There are probably better ways to project the 3D world onto a 2D surface. 
You can explore some of these options by looking at the help page for `coord_map` and `mapproj::mapproject`.
Let's use the [globular](https://en.wikipedia.org/wiki/Nicolosi_globular_projection) projection

```{r globularMap, exercise=TRUE, exercise.setup="preLoadMap"}
my_map + 
  coord_map('globular')
```

## Hurricane Map

For this next plot, we will use the built in `storms` dataset, and we will look particularly at the storms in $2016$. 

```{r HurricaneMap, exercise=TRUE, exercise.setup="preLoadMap"}
my_storms <- storms %>%
  filter(year == 2016, name %in% c('Matthew', "Nicole", "Karl", "Otto"))

my_map + 
  coord_map('globular') + 
  geom_path(data = my_storms, aes(x = long, y = lat), col = 'red') +
  # geom_circle(
  #   data = my_storms, 
  #   aes(x0 = long, y0 = lat, r = tropicalstorm_force_diameter/120),
  #   col = 'red'
  # ) +
  coord_map('globular') +  # results in distorted storm circles
  guides(color = "none") + 
  theme_bw() + 
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  ) + 
  facet_wrap(~name) + 
  ggtitle("Selected 2016 Tropical Storms and Hurricanes")
```

The variable `tropicalstorm_force_diameter` gives the diameter in nautical miles of the area experiencing tropical storm force winds. 
Because 1 degree of latitude corresponds to approximately 60 nautical miles, we can approximately display the impact area of each storm by adding circles with radius `tropicalstorm_force_diameter/120`. 
Note that these circles can be distorted based on our projection. 

## Table operations

Today we are going to focus on functions in the `dplyr` package, which is part of the `tidyverse`. 
The `dplyr` package aims to provide a function for each basic "verb" of data manipulation. 
Here are some common functions / "verbs" for data manipulation: 

* Rows: 
    + `filter()`: Select rows based on criteria on column values
    + `slice()`: choose rows based on locations.
    + `arrange()`: change the order of the rows.
* Columns: 
    + `select()`: Pick and choose columns to include in data.
    + `pull()`: Select a single column (similar to `data$column`).
    + `rename()`: Change the name of columns.
    + `mutate()`: Change the values of columns / creates new columns. 
* Groups (of rows):
    + `group_by()`: Creates a group of rows.
    + `summarise()` or `summarize()`: Calculate summary values for a given group. 

## `select()`, `filter()`, `mutate()` Examples

We are going to look at the `midwest` data again.

```{r}
head(midwest)
```

`select()` let's us select, or deselect, columns of interest

```{r midwestSelect1, exercise=TRUE}
midwest %>%
  select(popwhite, popwhite) %>%
  head()
```

```{r midwestSelect2, exercise=TRUE}
midwest %>%
  select(-PID, -poptotal) %>%
  head()
```

We can also select by condition: 

```{r midwestSelect3, exercise=TRUE}
midwest %>%
  select(where(is.character)) %>%
  head()
```

`filter()` allows us to select certain rows. 

```{r midwestFilter1, exercise=TRUE}
midwest %>%
  filter(state == 'MI') %>%
  head()
```

You can include more than one condition at a time: 

```{r midwestFilter2, exercise=TRUE}
midwest %>%
  filter(state == "MI", county == 'WASHTENAW')
```

**Exercise**: Subset the `midwest` data so that you only have data from the states of Wisconsin and Ohio, and only have the columns: `state`, `popdensity`, `inmetro`, and `percollege`. 

```{r midwestExercise1, exercise=TRUE}
midwest
```

## `filter` and `select` Solution

There are a few possible solutions. 
Here are two: 

```{r midwestSolution1}
midwest %>%
  filter(state %in% c("WI", "OH")) %>%
  select(state, popdensity, inmetro, percollege)
```

```{r midwestSolution2}
midwest %>%
  filter(state == "WI" | state == "OH") %>%
  select(state, popdensity, inmetro, percollege)
```

## Mutate
`mutate()` can be used to either create or modify a column in a dataset. 
We can pipe multiple functions together. 

```{r mutate1, exercise=TRUE}
midwest %>%
  filter(state == 'MI') %>%
  mutate(school = ifelse(county == "WASHTENAW", "GO BLUE!!!", "OTHER")) %>%
  mutate(county == tolower(county)) %>%
  select(state, county, school) %>%
  tail()
```

We can use `arrange()` to order the columns of the data. 

```{r arrange1, exercise=TRUE}
midwest %>%
  select(state, county) %>%
  arrange(desc(state), county) %>%
  head()
```

Another useful tool is `slice()`. This function allows us to select certain rows. There are a few versions of this that each give different results. See `?slice` for more details.

```{r slice1, exercise=TRUE}
midwest %>%
  select(state, county, poptotal) %>%
  filter(state == "MI") %>%
  slice_max(poptotal)
```

## `group_by()` and `summarize()` (1-2 slides)

`group_by()` helps us great groups. 
The following code will get the maximum population for each state in the dataset. 

```{r groupby1, exercise=TRUE}
midwest %>%
  select(state, county, poptotal) %>%
  group_by(state) %>%
  slice_max(poptotal)
```

`summarize()` (or equivalently, `summarise()`) is a function that's not particularly useful on it's own, but when combined with `group_by()` it can be very useful.
The following code counts the number of counties in each state

```{r summarize1, exercise=TRUE}
midwest %>%
  group_by(state) %>%
  summarize(n_counties = n()) %>%
  arrange(-n_counties)
```

## `group_by()` and `summarize()` Exercise.

**Exercise**

Use `group_by()` and `summarize()` to get the total population in each state. 
Then, arrange the states so that states with the smallest population comes first.

```{r groupExercise, exercise=TRUE}
midwest
```

**BONUS** Modify your previous solution to also calculate the percentage of the population that is White, Black, and Asian. 

## `group_by()` and `summarize()` Solution

```{r}
midwest %>%
  group_by(state) %>%
  summarize(state_population = sum(poptotal)) %>%
  arrange(state_population)
```

```{r}
midwest %>%
  group_by(state) %>%
  summarize(state_population = sum(poptotal),
            precent_black = sum(popblack) / state_population,
            percet_white = sum(popwhite) / state_population,
            percent_asian = sum(popasian) / state_population) %>%
  arrange(state_population)
```

Another possibility is to do all columns that start with *pop* at the same time: 

```{r}
midwest %>%
  group_by(state) %>%
  summarize(across(starts_with("pop"), ~ sum(.) / sum(poptotal)))
```

