---
title: "Stats 306: Final Review Session"
author: "Stats 306 GSIs"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

## Set up

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# This just checks if students need to install some packages that they might 
# not have.

```

```{r setup, eval=TRUE, include=TRUE, message=FALSE, echo=TRUE}
library(learnr)     # For interactive exercises
library(tidyverse)
```

```{r, include=FALSE, message=FALSE, echo=FALSE}
tutorial_options(exercise.reveal_solution=FALSE)
```

## Logistics

- Exam details: TODO.
- Today's outline: One hour of review, followed by one hour of time for questions.

## Lab 1 (Jesse / Yilei)

## Lab 2 (Victor)

### Plotting with `ggplot2`

A plot made using `ggplot2` has several components:

* The graph object itself (created using `ggplot()`)
* A set of *aesthetic* mappings connecting variables to visual properties
* Layers: collections of geometric elements (`geom_*()`) and statistical transformations (`stat_*()`)
* Scales: information on the range or composition of variables
* Coordinate systems: how the data are arranged spatially
* Facet: breaking a single plot into several similar plots for different subgroups
* Theme: all the other color and printing aspects of the plot

### Aesthetics and Aesthetic Mappings

An *aesthetic* is a visual property of the geometric elements in a plot.

```{r aesthetic_mappings_example}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy, color = drv))
```

Geometric elements represent observations in the dataset. In the plot above, the points are the geometric elements. Different geometric elements have different aesthetics. Some aesthetics of points are

* `x`, or x-coordinate
* `y`, or y-coordinate
* `color`
* `shape`
* `size`

An *aesthetic mapping* is an assignment of an aesthetic to a variable in the dataset. Aesthetic mappings are listed inside `aes()`. For example, in the previous plot, the aesthetic `x` was mapped to the variable `cty`.

### Geometric Elements (Geoms)

A *geometric element* or *geom* is a geometric object that is used in a plot to represent observations. A geom can represent

  * one observation;
  * a group of observations;
  * all of the observations.

A `ggplot2` plot consists of a base created by `ggplot()` and zero or more layers, each representing observations in a particular dataset using a particular geom.

### Statistical Transformations (Stats)

A *statistical transformation* or *stat* is an algorithm that computes from raw data values the values that will be plotted. The available stats are implemented in functions whose names begin with `stat_`. A stat is applied even when the raw values themselves are to be plotted. In that case, the stat that is applied is `stat_identity()`.

The dataset `diamonds` doesn't contain the minimum, median, and maximum for each cut:
```{r stat_summary_example2}
head(diamonds)
```

However, we can use `stat_summary()` to plot those summary statistics:
```{r stat_summary_example1}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

## Lab 3 (Jesse)

## Lab 4 (Yilei)

## Lab 5 (Benjamin)

## Lab 6 (Victor)

## Lab 7 (Jesse)

## Lab 8 (Yilei)

## Lab 9 (Benjamin)

## Lab 10 (Victor)

## Lab 11 (Benjamin)

## Lab 12 (Jesse)

## Lab 13 (Yilei)
