---
title: "Stats 306: Lab 9"
author: "Your Name"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

## Set up

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# This just checks if students need to install some packages that they might 
# not have.
if (!require(nycflights13)) install.packages("nycflights13", repos = "http://cran.us.r-project.org")
if (!packageVersion("learnr") == "0.11.1") install.packages("learnr", repos = "http://cran.us.r-project.org")
```

```{r setup, eval=TRUE, include=FALSE, message=FALSE, echo=FALSE}
library(learnr)     # For interactive exercises
library(tidyverse)  # For ggplot2, dplyr, readr
library(nycflights13)

tutorial_options(exercise.reveal_solution=FALSE)
```

```{r fake_setup, echo=TRUE, eval=FALSE}
library(learnr)     # For interactive exercises
library(tidyverse)  # For ggplot2, dplyr, readr
library(nycflights13)
```

## Logistics

-   Exam 2 is on Nov 11, more details to come.

-   Questions?

## Today's Content


- HW3 Review
- HW4 Review
- Strings
- EDA and Data Imports
- Tidy Data and Dates
- Table and Relational Data


## HW Review


## Strings

### Question

Recall that `stringr::str_c()` is used to concatenate strings.

```{r}
str_c(c("apple", "cucumber"), c("fruit", "vegetable"), sep = ": ")
```


Using the vector `base::LETTERS`, make a vector with 25 strings in which the first string is `"A precedes B"`, the second string is `"B precedes C"`, etc.

```{r str_c, exercise=TRUE}

```

```{r str_c-hint}
# Note that LETTERS[-1] contains B through Z
LETTERS[-1]
```

```{r str_c-solution}
str_c(LETTERS[-26], "precedes", LETTERS[-1], sep = " ")
```

## EDA and Data Imports


## Tidy Data and Dates


## Table and Relational Data

