---
title: "Stats 306: Lab 9"
author: "Your Name"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

## Set up

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# This just checks if students need to install some packages that they might 
# not have.
if (!require(nycflights13)) install.packages("nycflights13", repos = "http://cran.us.r-project.org")
if (!packageVersion("learnr") == "0.11.1") install.packages("learnr", repos = "http://cran.us.r-project.org")
```

```{r setup, eval=TRUE, include=FALSE, message=FALSE, echo=FALSE}
library(learnr)     # For interactive exercises
library(tidyverse)  # For ggplot2, dplyr, readr
library(nycflights13)

tutorial_options(exercise.reveal_solution=FALSE)
```

```{r fake_setup, echo=TRUE, eval=FALSE}
library(learnr)     # For interactive exercises
library(tidyverse)  # For ggplot2, dplyr, readr
library(nycflights13)
```

## Logistics

-   Exam 2 is on Nov 11, more details to come.

-   Questions?

## Today's Content


- HW3 Review
- HW4 Review
- Strings
- EDA and Data Imports
- Tidy Data and Dates
- Table and Relational Data


## HW 3 Review

## Question 1

Let’s look at some water quality data from the Backbay National Wildlife Refuge:

```{r}
bbwater <- read_csv("../data/BKB_WaterQualityData_2020084.csv")
```

## Q 1.1

Provide a short summary of the data. How many columns are there? What kinds of data do we have (what classes of R data)? How many columns are quantitative? How many are nominal? Are there any dates or times?

### Solution

```{r}
dim(bbwater)
```

We see 2371 observations and 17 columns. We can types from the `spec` function (though other methods of finding this information is also acceptable):

```{r}
spec(bbwater)
```

So we see character (string) and double (numeric) columns. We see that 7 columns are character, though it would probably be more accurate to say that only 5 of them are nominal as the `Read_Date` and `DateVerified` columns are ordinal or quantitative. So the remaining 17 - 5 = 12 are quantitative. In addition to the two dates previously listed, we also have a `Time` column.

## Q 1.2

Use `group_by` or `count` to find the number of observations at each `Site_Id`. Update the `bbwater` to remove any sites that have fewer than 5 observations.

### Solution

```{r}
count(bbwater, Site_Id)
```

We observe there are situations with fewer than 5 observations (it might be better to change “d” to “D”), which we remove

```{r}
bbwater <- group_by(bbwater, Site_Id) |> filter(n() > 5) |> ungroup()
count(bbwater, Site_Id)
```

## Q 1.3

Create an ECDF plot for the `Year` variable.

What is this plot telling you about measurements made before the late 1980s? Remember the refuge was started in 1938. Do any of these measurements seem out of place? Remove them from the `bbwater` table. Make a second ECDF after removing these probably erroneous entries.

### Solution

```{r}
ggplot(bbwater, aes(x = Year)) + stat_ecdf()
```

Going back to the description, “The refuge was established in 1938.” We see that there are at least some observations before this date, but not very many. Starting in the late 1980s, we see regular yearly readings with an slightl increase around the year 2000.

We’ll remove the pre-1935 observation (again, probably this typo for 19xx rather than 18xx.)

```{r}
bbwater <- filter(bbwater, Year > 1937)
ggplot(bbwater, aes(x = Year)) + stat_ecdf()
```

## Q 1.4

Here’s a little code that puts all of the numeric data into a table with two columns: `name` (which column) and `value` (the observed value for that variable).

```{r}
bbnumeric <- select(bbwater, where(is.numeric)) |> pivot_longer(everything())
```

Use this data to make a facet plot that shows a histogram plots for all the variables. Use the `scales = "free"` option to `facet_wrap`. Try a few different number of bins to pick a good number

After seeing the small multiples plot, select a variable to generate histogram where you carefully select the number of bins.

### Solution

```{r}
ggplot(bbnumeric, aes(x = value)) + geom_histogram(bins = 50) + facet_wrap(~ name, scales = "free")
```

```{r}
ggplot(bbwater, aes(x = `AirTemp (C)`)) + geom_histogram(bins = 100) # selected after a few tries, only needed to show 1
```

The `bins =` argument should have been set.

## Strings

### Question

Recall that `stringr::str_c()` is used to concatenate strings.

```{r}
str_c(c("apple", "cucumber"), c("fruit", "vegetable"), sep = ": ")
```


Using the vector `base::LETTERS`, make a vector with 25 strings in which the first string is `"A precedes B"`, the second string is `"B precedes C"`, etc.

```{r str_c, exercise=TRUE}

```

```{r str_c-hint}
# Note that LETTERS[-1] contains B through Z
LETTERS[-1]
```

```{r str_c-solution}
str_c(LETTERS[-26], "precedes", LETTERS[-1], sep = " ")
```

## EDA and Data Imports


## Tidy Data and Dates


## Table and Relational Data

