---
title: "Stats 306: Lab 10"
author: "Your Name"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

## Set up

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# This just checks if students need to install some packages that they might 
# not have.

if (!packageVersion("learnr") == "0.11.1") 
  install.packages("learnr", repos = "http://cran.us.r-project.org")
# For HW4 Q2, Q3
if (!require(Lahman)) 
  install.packages("Lahman", repos = "http://cran.us.r-project.org")
```

```{r setup, eval=TRUE, include=FALSE, message=FALSE, echo=FALSE}
library(Lahman) # For HW4 Q2, Q3
library(learnr)     # For interactive exercises
library(lubridate)
library(tidyverse)  # For dplyr, stringr

# For HW4 Q1
cancer_deaths <- read_csv("../../data/AH_Provisional_Cancer_Death_Counts_by_Month_and_Year__2020-2021.csv.gz")
cancer_deaths_2 <- select(cancer_deaths, - `Malignant neoplasms (C00-C97)`)

# For HW4 Q2
People <- mutate(People,
                 debut_d = ymd(debut),
                 finalGame_d = ymd(finalGame))
People <- mutate(People, career_length = finalGame_d - debut_d)

tutorial_options(exercise.reveal_solution=TRUE)
```

```{r fake_setup, echo=TRUE, eval=FALSE}
library(Lahman) # For HW4 Q2, Q3
library(learnr)     # For interactive exercises
library(tidyverse)  # For dplyr, stringr

# For HW4 Q1
cancer_deaths <- read_csv("../../data/AH_Provisional_Cancer_Death_Counts_by_Month_and_Year__2020-2021.csv.gz")
cancer_deaths_2 <- select(cancer_deaths, - `Malignant neoplasms (C00-C97)`)

# For HW4 Q2
People <- mutate(People,
                 debut_d = ymd(debut),
                 finalGame_d = ymd(finalGame))
People <- mutate(People, career_length = finalGame_d - debut_d)
```

## Logistics

- Quiz 4 is due by 11:59 PM on Nov 20.
- **Lab will be held on Nov 22.**
- Homework 5 is due by 11:59 PM on Dec 4.

## Today's Content

- HW4 Review
- Strings and Regular Expressions

## Q 1.2

```{r}
knitr::kable(head(cancer_deaths_2, 5))
```

Is `cancer_deaths2` in a tidy format (one row per observation, one column per measurement)? Use an appropriate `pivot_*` function to make this data more tidy (hint: see the help page for tidy select).

Call your table cancer_tidy. Use informative titles for the column names.

Display five rows of your table.

**Solution:** The clue that we are probably not in tidy form is that we are counting deaths in several different columns. It would be better to measure cancer types-demographic groups per month as the unit so that deaths can be in a common column.
```{r}
cancer_tidy <- pivot_longer(cancer_deaths_2, contains("malignant"), values_to = "deaths", names_to = "cancer_type")

knitr::kable(head(cancer_tidy, 5))
```

## Q 1.4

```{r}
cancer_tidy_sparse <- filter(cancer_tidy, deaths > 0)
```

Use the sparse version to answer the question, what were the top 5 forms of cancer by total deaths?

**Solution:**
```{r}
group_by(cancer_tidy_sparse, cancer_type) |> 
  summarize(sum(deaths)) |> 
  arrange(desc(`sum(deaths)`)) |> head(5) |>
  knitr::kable()
```

## Q 1.5

Grouping by “Age Group” and cancer type, find the single most deadly cancer (by total deaths) in each group. Do age groups vary in the most deadly form of cancer or all they all the same?

**Solution:**
```{r}
group_by(cancer_tidy_sparse, `Age Group`, cancer_type) |>
  summarize(d = sum(deaths)) |>
  arrange(desc(d)) |>
  summarize(first(cancer_type)) |>
  knitr::kable()
```

We observe variation in the type of cancer per age group. For youth, brain and CNS. For the next 20 years, a mix of other, less common types. Then breast cancer hits women particularly hard. For the older generations, lung cancer and related cancers.

## Q 2.3

Write a function that takes a date and returns the decade. For example, on the input `c("1891-07-10", "1997-10-01")` should return `c(1890, 1990)`. (Hint see the `round` function and its use of negative rounding.)

Compute the decade of each player’s debut and plot a series of boxplots or violin plots of `career_length` by decade of debut. What trends do you notice? How do you explain the decrease in career length in recent years? (hint: what did you see in 2.1?)

**Solution:**

The hint rounded up and down, so not exactly what was asked for in the text. We can do it both ways:

```{r}
decade_floor <- function(ds) {
  y <- year(ds)
  floor(y / 10) * 10
}
decade_round <- function(ds) {
  y <- year(ds)
  round(y, -1)
}
```

```{r}
ggplot(People, aes(x = factor(decade_round(debut_d)), y = career_length)) +
  geom_boxplot()
```

```{r}
ggplot(People, aes(x = factor(decade_floor(debut_d)), y = career_length)) +
  geom_boxplot()
```

It looks like career lengths are decreasing in the most recent decade, but this could be the product of the censoring we observed in the previous part of the problem, i.e., of not yet observing how long current players will play in the future.

## Q 2.4

Filtering out the players with `NAs` for `debut_d` or `finalGame_d`. Which player has had the longest career? (Display the name along with the duration in years.) Look up this player on Wikipedia to learn how he had such a long career. Was he playing the entire time? With which other player did he perform comedy routines?

For players continuing to play, we have a “censored” career length as we have not yet observed their total career, but such is the nature of time. Of players that have the maximum `finalGame_d`, which player has had the longest career? (Display the name and duration in years using the `time_length` function)

**Solution:**

Overall:

```{r}
filter(People, !is.na(career_length), 
       career_length == max(career_length, na.rm = TRUE)) |>
  select(playerID, nameFirst, nameLast, debut_d, finalGame_d, career_length) |>
  mutate(career_length = time_length(career_length, "years"))
```

Only pinch hit later in life. Comedy partner, Al Schact.

Currently playing:

```{r}
filter(People, !is.na(career_length),
       finalGame_d == max(finalGame_d, na.rm = TRUE)) |>
  select(playerID, nameFirst, nameLast, debut_d, finalGame_d, career_length) |>
  mutate(career_length = time_length(career_length, "years")) |>
  arrange(desc(career_length)) |> head(1)
```

## Q 3.1

Review the documentation for the `People` table and the `Fielding` table. Answer the following:

What is the unit (row) for the `People` table?
What is the unit (row) for the `Fielding` table? (Hint: see the code below)
What columns are common to both both tables?

```{r}
group_by(Fielding, playerID, yearID) |> summarize(years = n()) |> ungroup() |> summarize(sum(years > 1))
```

```{r}
group_by(Fielding, playerID, yearID, stint) |> summarize(stints = n()) |> ungroup() |> summarize(sum(stints > 1))
```

```{r}
group_by(Fielding, playerID, yearID, stint, POS) |> summarize(positions = n()) |> ungroup() |> summarize(sum(positions > 1))
```

**Solution:**

The `People` table is one row per person. This can be obtained either from the documentation or from code like this:

```{r}
count(People, playerID) |> summarize(all(n < 1))
```

For the `Fielding` table, we have a record for each combination of player, year, stint (a segment of year) and position.

For common column names, we can use a set operation:

```{r}
intersect(colnames(People), colnames(Fielding))
```

## Q 3.2

Using a `inner_join`, find the names of the top 5 players for total double plays. (Hint: you will need to do some grouping and summarizing before your join.)

**Solution:**

```{r}
group_by(Fielding, playerID) |> summarize(DP = sum(DP)) |>
  inner_join(People, by = "playerID") |>
  arrange(desc(DP)) |> head(5) |>
  select(nameFirst, nameLast, DP)
```

## Q 3.3

Review the `Managers` and `AwardsManagers` table.

Use a `left_join` to determine what proportion of managers won an award in each year. Graph this as a line plot. What do you notice about the proportion of managers winning awards?

Note: you will need to do some grouping first. What is this code snippet telling you:

```{r}
group_by(Managers, playerID, yearID, teamID) |> summarize(n()) |> nrow()
```

```{r}
nrow(Managers)
```

**Solution:**

```{r}
group_by(Managers, playerID, yearID, teamID) |> summarize(teams = n()) |>
  left_join(AwardsManagers, by = c("playerID", "yearID")) |>
  group_by(playerID, yearID) |>
  summarize(a = sum(!is.na(awardID))) |>
  group_by(yearID) |> summarize(proportion_winning_award = mean(a)) ->
  year_awards
```

```{r}
ggplot(year_awards, aes(x = yearID, y = proportion_winning_award)) +
  geom_line()
```

We see a decreasing trend but then a big jump. There must have been a big increase in the number of awards given in the 1990s.

## Strings

Recall that a **string** is a sequence of characters.

Below is an example of a string as it would appear on paper.
```
backyard
```

Strings in R work differently than they do on paper. In R, a string **must either begin and end with a double quote `"` or begin and end with a single quote `'`.**
```{r}
"backyard"
'backyard'
```

**Question:** What would happen if you entered `"backyard"` without the quotes and why?
```{r no_quotes, exercise=TRUE}

```

```{r no_quotes-solution}
backyard
# Without the quotes, the interpreter would assume that backyard is an
# object that we want to print. It wouldn't be able to find backyard because
# it doesn't exist and would then throw an error.
```

`writeLines()` takes the R version of a string and shows us the paper version.
```{r, comment=NA}
writeLines("backyard")
```

Certain characters are inserted in strings using **escape sequences**. An escape sequence starts with a backslash `\`. For example, the escape sequence for `tab` is `\t`. The backslash combines with the next character to form an escape sequence. See the help page on `Quotes` for the list of valid escape sequences in R. This is how a tab would be inserted in the previous string.
```{r}
"back\tyard"
```

This may be more natural, but it's worse because it's ambiguous.
```{r}
"back    yard"
```

To include a backslash, use the escape sequence `\\`.
```{r}
"back\\yard"
```

Another commonly used escape sequence is the newline escape sequence. A newline represents the start of a new line.
```{r}
"back\nyard"
```

**Question:** What would happen if you tried to insert a tab, or a backslash, or a newline in the middle of `"backyard"` without using an escape sequence and why? 
```{r escape_seqs, exercise=TRUE}

```

```{r escape_seqs-solution}
# Inserting a tab by hitting the tab key doesn't work. In the console, it would
# bring up the autocomplete box.
# Inserting a backslash with just one backslash causes \y to be interpreted as
# an escape sequence, which causes an error as \y isn't a valid escape
# sequence.
# Inserting a newline by hitting the return key seems to work, but R seems to
# think that the code has a mistake
```

Here are the outputs of `writeLines()` for the three modifications of `"backyard"`. They look the way they would on paper.
```{r, comment=NA}
writeLines("back\tyard")
writeLines("back\\yard")
writeLines("back\nyard")
```

Since quotes are used to mark the start and end of an R string, an escape sequence must be used to put a quote in the middle of an R string if it's of the same type as the start and end quotes.
```{r, comment=NA}
"back\"yard" 
writeLines("back\"yard")
```

**Question:** What happens if you don't escape and why?
```{r no_escaping, exercise=TRUE}

```

```{r no_escaping-solution}
# The interpreter sees the middle quote and thinks that the string has ended,
# but then sees another character, which doesn't make any sense, so it throws
# an error
"back"yard"
```

However, if the quote is of the other type, escaping isn't necessary.
```{r, comment=NA}
"back'yard"
writeLines("back'yard")
```

## Strings: More Functions

We've talked about several functions for working with strings: `str_c()`, `str_length()`, `str_sub()`, `str_split()`, etc. Some additional useful string functions are described below.

`strrep()` is used to repeat a string. `strrep()` is in the `base` package unlike the other string functions we've covered.
```{r}
strrep(c("apple", "banana", "cherry"), 1:3)
```

`str_interp()` makes a string from a template by inserting the values of variables. This can be very useful when making titles for plots. Note that a variable must be enclosed in braces and a dollar sign must be put before the left brace.
```{r}
x <- 1
y <- 2
z <- 3
str_interp("${x} + ${y} = ${z}")
```

### Exercise

Without hard-coding, recreate the ASCII art below. There are 10 lines, each of which has 8 hyphens in the middle.
```
|----------|
|----------|
|----------|
|----------|
|----------|
|----------|
|----------|
|----------|
|----------|
|----------|
```

```{r strrep, exercise=TRUE}

```

```{r strrep-hint}
# Try using strrep() and writeLines() in a for loop
```

```{r strrep-solution}
for (n in 1:10) {
  line <- str_c("|", strrep("-", 10), "|")
  writeLines(line)
}
```

### Exercise

Recall that `iris` is a data frame with measurements of petal and sepal length and width for a few flower species. `make_scatterplot()` takes a species and two measurement variables and makes a scatterplot of the variables for that species. Complete `make_scatterplot()`. If the species is `virginica`, the `x` variable is `Sepal.Length`, and the `y` variable is `Sepal.Width`, then the title of the plot should be `"Sepal.Width vs. Sepal.Length for virginica"`
```{r str_interp, exercise=TRUE}
make_scatterplot <- function(species, x_var, y_var) {
  title <- # Fill in
  iris %>%
    # Fill in - only use data for species
    ggplot(aes(!!as.name(x_var), !!as.name(y_var))) +
    # Fill in - choose the right geom
    # Fill in - set the title
    theme_bw()
}
make_scatterplot("virginica", "Sepal.Length", "Sepal.Width")
```

```{r str_interp-solution}
make_plot <- function(species, x_var, y_var) {
  title <- str_interp("${y_var} vs. ${x_var} for ${species}")
  iris %>%
    filter(Species == species) %>%
    ggplot(aes(!!as.name(x_var), !!as.name(y_var))) +
    geom_point() +
    ggtitle(title) +
    theme_bw()
}
make_plot("virginica", "Sepal.Length", "Sepal.Width")
```

### Optional Challenge Exercise

Without hard-coding, recreate the ASCII art below. There are 8 lines, each of which has 7 hyphens in the middle.
```
|\-------|
|-\------|
|--\-----|
|---\----|
|----\---|
|-----\--|
|------\-|
|-------\|
```

```{r challenge, exercise=TRUE}

```

```{r challenge-solution}
for (n in 0:7) {
  line <- str_c("|", strrep("-", n), "\\", strrep("-", 7 - n), "|") 
  writeLines(line)
}
```


## Regular Expressions

A **regular expression** is a string specifying a pattern that other strings may or may not match. Regular expressions can be used to

  - find strings matching a pattern
  - modify substrings matching a pattern
  - delete substrings matching a pattern
  
An example use of regular expressions is validation of email addresses. If a company wants to check whether a customer's email address is valid, they could check whether it matches a regular expression like [this one](https://stackoverflow.com/a/201378):

```
(?:[a-z0-9!#$%&'*+/=?^_`{|}~-]+(?:\.[a-z0-9!#$%&'*+/=?^_`{|}~-]+)*|"(?:[\x01-\x08\x0b\x0c\x0e-\x1f\x21\x23-\x5b\x5d-\x7f]|\\[\x01-\x09\x0b\x0c\x0e-\x7f])*")@(?:(?:[a-z0-9](?:[a-z0-9-]*[a-z0-9])?\.)+[a-z0-9](?:[a-z0-9-]*[a-z0-9])?|\[(?:(?:(2(5[0-5]|[0-4][0-9])|1[0-9][0-9]|[1-9]?[0-9]))\.){3}(?:(2(5[0-5]|[0-4][0-9])|1[0-9][0-9]|[1-9]?[0-9])|[a-z0-9-]*[a-z0-9]:(?:[\x01-\x08\x0b\x0c\x0e-\x1f\x21-\x5a\x53-\x7f]|\\[\x01-\x09\x0b\x0c\x0e-\x7f])+)\])
```

### Metacharacters

Several characters have special meanings inside regular expressions. They are called **metacharacters**. These are the metacharacters:

`. \ | ( ) [ ] ^ $ { } * + ?`

For example, `.` matches any character.

If you do not want a metacharacter to have its special meaning, you need to prepend it with **two** backslashes.
```{r}
str_view_all(c("$2.50", "2 dollars"), "\\.")
```

**Question:** What happens if you use only one backslash or no backslashes and why?
```{r too_few_backslashes, exercise=TRUE}

```

```{r too_few_backslashes-solution}
# Both strings match since . matches any character
str_view_all(c("$2.50", "2 dollars"), ".")
# Causes an error since \. isn't a valid escape sequence
str_view_all(c("$2.50", "2 dollars"), "\.")
```

An alternative is to use `fixed()`.
```{r}
str_view_all(c("$2.50", "2 dollars"), fixed("."))
```

### Character Classes

- `.` matches any character
- `[abcd]` matches any one of the characters between the brackets
- `[^abcd]` matches any character *not* between the brackets (excluding `^`)
- `[a-d]` matches any character in the specified range; it's the same as `[abcd]`

An example:
```{r}
str_view_all(c("$2.50", "2 dollars"), "[a-zA-Z]")
```

An alternative way to make the matching case-insensitive:
```{r}
str_view_all(c("$2.50", "2 dollars"), regex("[a-z]", ignore_case = TRUE))
```


### Shorthand

- `\w` matches a "word" character, equivalent to `[0-9A-Za-z_]`, i.e., digits, letters, and the underscore
- `\s` matches any whitespace character, including tabs and newlines
- `\d` matches digits, equivalent to `[0-9]`
- `\W`, `\S`, and `\D` match the opposite of the lower-case versions

In the example below, we check which string contains whitespace. Note that **two** backslashes need to be used. Why?
```{r}
str_view_all(c("$2.50", "2 dollars"), "\\s")
```

### Grouping

- `()` are used to group patterns together. This can be used to extract matches individually.
- `\1` refers to the match to the first group, `\2` refers to the match to the second group, etc.

In this example, we check which string contains a repeated letter.
```{r}
str_view_all(c("$2.50", "2 dollars"), "([a-z])\\1")
```

### Operators

- `|` is the OR operator and allows matches of either side
- `{}` describes how many times the preceeding character or group must occur:
  - `{m}` must occur exactly `m` times
  - `{m,n}` must occur between `m` and `n` times, inclusive
  - `{m,}` Must occur at least `m` times
- `*` means the preceeding character can appear zero or more times, equivalent to `{0,}`
- `+` means the preceeding character must appear one or more times, equivalent to `{1,}`
- `?` means the preceeding character can appear zero or one time, equivalent to `{0,1}`

Which string has two digits in a row?
```{r}
str_view_all(c("$2.50", "2 dollars"), "[0-9]{2}")
```

### Anchors

- `^` matches the start of a string (or line)
- `$` matches the end of a string (or line)
- `\b` matches a word boundary
- `\B` matches not word boundary

Which string starts with a dollar sign? Note that `$` is escaped with **two** backslashes.
```{r}
str_view_all(c("$2.50", "2 dollars"), "^\\$")
```

## Q 4.2

Use the corpus of common words in `stringr::words`, create regular expressions that find all words that:

- Start with `“y”`.
- End with `“x”`
- Are exactly three letters long. (Don’t cheat by using `str_length()`!)
- Have seven letters or more.

For each, provide the size of each set.

**Solution:**

```{r}
f <- function(regex) { str_detect(stringr::words, regex) |> sum() }

tribble(
   ~ goal, ~ count,
   'Start with "y"', f("^y"),
   'End with "x"', f("x$"),
   'three letters long', f("^[a-zA-Z]{3}$"),
   '7+ letters', f("^[a-zA-Z]{7,}"),
)
```

## Q 4.3

Continuing with the common words list, create regular expressions to find all words that:

- Start with a vowel.
- That only contain consonants. (Hint: thinking about matching “not”-vowels.)
- End with “ed”, but not with “eed”.
- End with “ing” or “ise”.

**Solutions:**

Two solutions for the ones relating to vowels in case you put “y” in that group

```{r}
tribble(
   ~ goal, ~ count,
   "vowel", f("^[aeiouAEIOU]"),
   "vowels and y", f("^[aeiouyAEIOUY]"),
   "consontants", f("^[^aeiouAEIOU]+$"),
   "consontants not y", f("^[^aeiouyAEIOUY]+$"),
   "ed not eed", f("[^e]ed$"),
   "ing or ise", f("(ing|ise)$")
)
```

## Q 4.4

Using a regular expression and `select`, find the columns in `cancer_deaths` that only refer to a single type of cancer.

**Solution:**

```{r}
select(cancer_deaths, matches("\\(C\\d\\d\\)")) |> colnames()
```

Alternative if “singular” is taken to exclude C22.
```{r}
select(cancer_deaths, matches("neoplasm[^s]")) |> colnames()
```

## Regular Expression Exercises

The exercises below use this paragraph.

```{r}
baseball <- "According to Baseball Reference’s wins above average, The Red Sox had the best 
outfield in baseball— one-tenth of a win ahead of the Milwaukee Brewers, 11.5 to 11.4. And 
that’s despite, I’d argue, the two best position players in the NL this year (Christian 
Yelich and Lorenzo Cain) being Brewers outfielders. More importantly, the distance from 
Boston and Milwaukee to the third-place Yankees is about five wins. Two-thirds of the Los 
Angeles Angels’ outfield is Mike Trout (the best player in baseball) and Justin Upton (a 
four-time All-Star who hit 30 home runs and posted a 122 OPS+ and .348 wOba this year), 
and in order to get to 11.5 WAA, the Angels’ outfield would have had to replace right 
fielder Kole Calhoun with one of the three best outfielders in baseball this year by WAA."
```

The exercises require you to extract matches using `str_extract_all()`

### Exercise 1

Extract all the capitalized words in the paragraph.

```{r baseball1, exercise=TRUE}

```

```{r baseball1-hint}
# Try using \b
```

```{r baseball1-solution}
# `\b` looks for a word boundary (not just the beginning of the text snippet!)
# `[A-Z]` matches a single capitalized letter
# `[a-z]` matches a single lowercase letter
# `+` means we match arbitrarily many lowercase letters
str_extract_all(baseball, "\\b[A-Z][a-z]+")
```

### Exercise 2

Extract all the numbers.

```{r baseball2, exercise=TRUE}

```

```{r baseball2-solution}
str_extract_all(baseball, "\\.?\\d+\\.?\\d*")
```

### Exercise 3

Extract all the hyphenated words.

```{r baseball3, exercise=TRUE}

```

```{r baseball3-solution}
str_extract_all(baseball, regex("[a-z]+-[a-z]+", ignore_case = TRUE))
```

### Exercise 4

Extract all words with two consecutive vowels.

```{r baseball4, exercise=TRUE}

```

```{r baseball4-solution}
str_extract_all(baseball, regex("[a-z]*[aeiou]{2}[a-z]*", ignore_case = TRUE))
```

### Exercise 5

Extract all words with a repeated letter.

```{r baseball5, exercise=TRUE}

```

```{r baseball5-solution}
str_extract_all(baseball, regex("[a-z]*([a-z])\\1[a-z]*", ignore_case = TRUE))
```